### 2015-11-13 proposal/proposal.tex
Init cabal and separate proposal files to dir proposal

\documentclass{article}
\usepackage{natbib}

\title{\bf Final Project Proposal}
\author{Your Name}
\date{Due date: 16 November 2015 at 11:59 PM Central time}

\begin{document}
\maketitle

% Lines that begin with % are comments.

\section{Motivation}
\label{sec:motivation}

Write an introductory section that motivates a particular problem you wish to solve.

This section should (at a minimum) fill the remainder of this page.

\section{Related Work}

Briefly survey the most salient prior work that relates to the problem you wish to solve. This section should cite relevant sources. This section should be at least one to two paragraphs in length.

As we saw is Section \ref{sec:motivation}, ...
%
% For example: 
%

\citet{Weaver_memo_1949} proposed that machine translation \ldots
%
The square root of 4 is 2 \citep{1952_06_17_BarHillel}.

\section{Proposal}

Write your proposal here.

\bibliographystyle{apalike}
\bibliography{your_bibliography}

\end{document}

### 2015-11-14 src/Main.hs
Add src dir

@@ -1,0 +1,6 @@
+module Main where
+
+data POS = Noun
+         | Verb
+
+main = putStrLn "Hi"

### 2015-11-14 src/Main.hs
Add ADT for representing a POS

@@ -1,6 +1,18 @@
 module Main where
 
 data POS = Noun
+         | Adj
+         | Adv
          | Verb
+         | Pron
+         | Conj
+         | Det
+         | Postp
+         | Ques
+         | Interj
+         | Num
+         | Dup
+         | Punc
+         deriving (Eq, Show,)
 
 main = putStrLn "Hi"

### 2015-11-14 src/Main.hs
Required imports

@@ -1,5 +1,10 @@
 module Main where
 
+import Data.HMM
+import Control.Monad
+import Data.Array
+import System.IO
+
 data POS = Noun
          | Adj
          | Adv
@@ -13,6 +18,6 @@ data POS = Noun
          | Num
          | Dup
          | Punc
-         deriving (Eq, Show,)
+         deriving (Eq, Show)
 
 main = putStrLn "Hi"

### 2015-11-15 src/Main.hs
Modify config to include vector

@@ -2,8 +2,8 @@ module Main where
 
 import Data.HMM
 import Control.Monad
-import Data.Array
 import System.IO
+import qualified Data.Vector.Storable as VS
 
 data POS = Noun
          | Adj

### 2015-11-15 src/Main.hs
Get XML files from the treebank

@@ -1,9 +1,13 @@
+{-# LANGUAGE Arrows #-}
+
 module Main where
 
 import Data.HMM
 import Control.Monad
 import System.IO
+import System.Directory
 import qualified Data.Vector.Storable as VS
+import qualified Text.XML.HXT.Core as HXT
 
 data POS = Noun
          | Adj
@@ -20,4 +24,7 @@ data POS = Noun
          | Punc
          deriving (Eq, Show)
 
-main = putStrLn "Hi"
+main :: IO ()
+main = do
+  files <- getDirectoryContents "tb_corrected"
+  print $ files !! 2

### 2015-11-15 src/Main.hs
Preliminary XML parsing

@@ -1,13 +1,13 @@
-{-# LANGUAGE Arrows #-}
+{-# LANGUAGE Arrows, NoMonomorphismRestriction #-}
 
 module Main where
 
-import Data.HMM
-import Control.Monad
-import System.IO
+-- import Data.HMM
+import System.IO hiding (putStr)
 import System.Directory
-import qualified Data.Vector.Storable as VS
-import qualified Text.XML.HXT.Core as HXT
+-- import qualified Data.Vector.Storable as VS
+import Text.XML.HXT.Core
+import Prelude hiding (words, putStr)
 
 data POS = Noun
          | Adj
@@ -24,7 +24,20 @@ data POS = Noun
          | Punc
          deriving (Eq, Show)
 
+parseFile :: FilePath -> IOStateArrow a b XmlTree
+parseFile f = readDocument [ withValidate no
+                           , withRemoveWS yes
+                           ] f
+
+words = deep (isElem >>> hasName "S") >>>
+  proc x -> do
+    word <- getText <<< getChildren <<< deep (hasName "W") -< x
+    returnA -< word
+
+
 main :: IO ()
 main = do
-  files <- getDirectoryContents "tb_corrected"
-  print $ files !! 2
+  files <- getDirectoryContents "tb_uni"
+  let file = "tb_uni/" ++ files !! 2
+  ws <- runX (readDocument [withValidate no] file >>> words)
+  putStrLn $ unwords ws

### 2015-11-15 proposal/proposal.tex
Begin working on motivation

@@
 \usepackage{natbib}
+\usepackage{hyperref}
 
-\title{\bf Final Project Proposal}
+\title{\bf An HMM-based part-of-speech tagger for Turkish}
 \author{Your Name}
@@
 
-% Lines that begin with % are comments.
-
 \section{Motivation}
 \label{sec:motivation}
-
-Write an introductory section that motivates a particular problem you wish to solve.
-
-This section should (at a minimum) fill the remainder of this page.
+My main motivation for undertaking this project is to understand the main
+principles underlying POS-tagging; due to time and resource constraints, it is
+not possible to build a competitive POS-tagger for Turkish without doing
+morphological analysis hence this project. Nevertheless, I expect my POS-tagger
+to work with a decent accuracy that is good enough to be a starting point for
+further development. Moreover, \citet{Korkut2015} began the development of a
+Turkish NLP framework implemented in the programming language Haskell. I intend to incorporate the
+POS-tagger to the Guguk project, hence contributing towards the existence of a
+Turkish NLP-framework in Haskell.
 
@@
 
-Briefly survey the most salient prior work that relates to the problem you wish to solve. This section should cite relevant sources. This section should be at least one to two paragraphs in length.
+% Briefly survey the most salient prior work that relates to the problem you wish
+% to solve. This section should cite relevant sources. This section should be at
+% least one to two paragraphs in length.
+The most important advancement in the computational linguistics of Turkish is
+the NLP framework ``Zemberek'' by \citet{akin2007zemberek}. Although it started
+out as an NLP tool for the Turkish language, Zemberek moved on to become a
+framework for dealing with all Turkic languages
 

### 2015-11-16 src/Main.hs
Strip tags and construct type using the tag string

@@ -22,22 +22,32 @@ data POS = Noun
          | Num
          | Dup
          | Punc
-         deriving (Eq, Show)
+         deriving (Eq, Show, Read)
 
 parseFile :: FilePath -> IOStateArrow a b XmlTree
 parseFile f = readDocument [ withValidate no
                            , withRemoveWS yes
                            ] f
 
-words = deep (isElem >>> hasName "S") >>>
-  proc x -> do
-    word <- getText <<< getChildren <<< deep (hasName "W") -< x
-    returnA -< word
+getPOS :: String -> POS
+getPOS x = read (takeWhile (\n -> n /= '+' && n /= '"') $ tail $ dropWhile (/= '+') x)
 
+atTag tag = deep (isElem >>> hasName tag)
+
+words = atTag "S" >>>
+  proc x -> do
+    w    <- (atTag "W") -< x
+    word <- getText <<< getChildren -< w
+    tag  <- getAttrValue "IG" -< w
+    returnA -< (word, tag)
+
+printTuple :: (String, String) -> IO ()
+printTuple (a, b)  = putStrLn (a ++ " " ++ (show $ getPOS b))
 
 main :: IO ()
 main = do
   files <- getDirectoryContents "tb_uni"
   let file = "tb_uni/" ++ files !! 2
   ws <- runX (readDocument [withValidate no] file >>> words)
-  putStrLn $ unwords ws
+  _  <- mapM printTuple ws
+  return ()

### 2015-11-16 proposal/proposal.tex
Update proposal

@@
 \title{\bf An HMM-based part-of-speech tagger for Turkish}
-\author{Your Name}
-\date{Due date: 16 November 2015 at 11:59 PM Central time}
+\author{Ayberk Tosun\\\texttt{tosun2@illinois.edu}}
+\date{}
 
@@
 further development. Moreover, \citet{Korkut2015} began the development of a
-Turkish NLP framework implemented in the programming language Haskell. I intend to incorporate the
-POS-tagger to the Guguk project, hence contributing towards the existence of a
-Turkish NLP-framework in Haskell.
+Turkish NLP framework implemented in the programming language Haskell. I intend
+to incorporate the POS-tagger to the Guguk project, hence contributing towards
+the existence of a Turkish NLP-framework in Haskell.
 
@@
 out as an NLP tool for the Turkish language, Zemberek moved on to become a
-framework for dealing with all Turkic languages
+a tool for processing all Turkic languages. Zemberek has been immensely useful
+for linguistic research on Turkish. Yet, it is undeniable that once-popular
+Java does not provide an ideal environment for writing NLP software.
+
+The widely cited \citet{oflazer1994tagging} outlines the implementation of a
+tool for both morpheme glossing and POS tagging, that has ``98-99 \%'' accuracy. This
+tool is based on a morphological specification of Turkish and makes use of a
+morphological processor called ``PC-Kimmo'' \citep{antworth1991pc}.
 
@@
 %
-
-\citet{Weaver_memo_1949} proposed that machine translation \ldots
-%
-The square root of 4 is 2 \citep{1952_06_17_BarHillel}.
 

### 2015-11-16 src/Main.hs
Add type annotations for top-levels

@@ -9,6 +9,8 @@ import System.Directory
 import Text.XML.HXT.Core
 import Prelude hiding (words, putStr)
 
+type TaggedWord = (String, POS)
+
 data POS = Noun
          | Adj
          | Adv
@@ -29,25 +31,27 @@ parseFile f = readDocument [ withValidate no
                            , withRemoveWS yes
                            ] f
 
-getPOS :: String -> POS
-getPOS x = read (takeWhile (\n -> n /= '+' && n /= '"') $ tail $ dropWhile (/= '+') x)
+extractPOS :: String -> POS
+extractPOS x = read (takeWhile (\n -> n /= '+' && n /= '"') $ tail $ dropWhile (/= '+') x)
 
+atTag :: ArrowXml a => String -> a XmlTree XmlTree
 atTag tag = deep (isElem >>> hasName tag)
 
+words :: ArrowXml cat => cat XmlTree (String, String)
 words = atTag "S" >>>
   proc x -> do
-    w    <- (atTag "W") -< x
-    word <- getText <<< getChildren -< w
-    tag  <- getAttrValue "IG" -< w
-    returnA -< (word, tag)
+    w       <- (atTag "W") -< x
+    word    <- getText <<< getChildren -< w
+    tagStr  <- getAttrValue "IG" -< w
+    returnA -< (word, tagStr)
 
 printTuple :: (String, String) -> IO ()
-printTuple (a, b)  = putStrLn (a ++ " " ++ (show $ getPOS b))
+printTuple (a, b)  = putStrLn (a ++ " " ++ (show $ extractPOS b))
 
 main :: IO ()
 main = do
   files <- getDirectoryContents "tb_uni"
-  let file = "tb_uni/" ++ files !! 2
+  let file = "tb_uni/" ++ files !! 3
   ws <- runX (readDocument [withValidate no] file >>> words)
   _  <- mapM printTuple ws
   return ()

### 2015-11-16 src/Main.hs
Refactor

@@ -32,7 +32,9 @@ parseFile f = readDocument [ withValidate no
                            ] f
 
 extractPOS :: String -> POS
-extractPOS x = read (takeWhile (\n -> n /= '+' && n /= '"') $ tail $ dropWhile (/= '+') x)
+extractPOS x = read $ takeWhile notPlusNorQuote . tail $ dropWhile notPlus x
+  where notPlus = (/= '+')
+        notPlusNorQuote = \n -> notPlus n && n /= '"'
 
 atTag :: ArrowXml a => String -> a XmlTree XmlTree
 atTag tag = deep (isElem >>> hasName tag)

### 2015-11-16 src/Main.hs
Add comments

@@ -11,6 +11,8 @@ import Prelude hiding (words, putStr)
 
 type TaggedWord = (String, POS)
 
+-- | These are the possible parts-of-speech described in the
+-- | METU-Sabanci treebank paper.
 data POS = Noun
          | Adj
          | Adv
@@ -26,11 +28,6 @@ data POS = Noun
          | Punc
          deriving (Eq, Show, Read)
 
-parseFile :: FilePath -> IOStateArrow a b XmlTree
-parseFile f = readDocument [ withValidate no
-                           , withRemoveWS yes
-                           ] f
-
 extractPOS :: String -> POS
 extractPOS x = read $ takeWhile notPlusNorQuote . tail $ dropWhile notPlus x
   where notPlus = (/= '+')
@@ -39,6 +36,11 @@ extractPOS x = read $ takeWhile notPlusNorQuote . tail $ dropWhile notPlus x
 atTag :: ArrowXml a => String -> a XmlTree XmlTree
 atTag tag = deep (isElem >>> hasName tag)
 
+-- | For all children of <S> (i.e., the words) return a tuple containing
+-- | the word and the string containing part-of-speech and morpheme
+-- | glossing of the word. We are primarily interested in the part-of-speech
+-- | contained by this `tagStr` so we will eventually extract it out using
+-- | `extractPOS`.
 words :: ArrowXml cat => cat XmlTree (String, String)
 words = atTag "S" >>>
   proc x -> do

### 2015-11-16 proposal/proposal.tex
Update proposal

@@
 principles underlying POS-tagging; due to time and resource constraints, it is
-not possible to build a competitive POS-tagger for Turkish without doing
-morphological analysis hence this project. Nevertheless, I expect my POS-tagger
-to work with a decent accuracy that is good enough to be a starting point for
+not possible for me to build a state-of-the-art POS-tagger for
+Turkish. Nevertheless, I expect my POS-tagger
+to work with a decent accuracy that is enough to be a starting point for
 further development. Moreover, \citet{Korkut2015} began the development of a
-Turkish NLP framework implemented in the programming language Haskell. I intend
-to incorporate the POS-tagger to the Guguk project, hence contributing towards
+Turkish NLP framework named ``Guguk'' implemented in the programming language Haskell. I intend
+to incorporate this project to the Guguk project, hence contributing towards
 the existence of a Turkish NLP-framework in Haskell.
@@
 \section{Related Work}
+\label{sec:related_work}
 
@@
 a tool for processing all Turkic languages. Zemberek has been immensely useful
-for linguistic research on Turkish. Yet, it is undeniable that once-popular
-Java does not provide an ideal environment for writing NLP software.
+for linguistic research on Turkish.
 
@@
 
-As we saw is Section \ref{sec:motivation}, ...
-%
-% For example: 
-%
-
+Most of the work relating to the computational processing of Turkish deal with
+Turkish morphology since Turkish is a morphologically rich language analyzing
+the morphology seems to be the more salient problem e
 \section{Proposal}
 
-Write your proposal here.
+I will be implementing a part-of-speech tagger for the Turkish language. As
+mention in section~\ref{sec:related_work} I will be mostly focusing on parts-of-speech
+of words and will not be accounting for morphology. There are problems with this
+approach but this project aims to be a starting point for further research.
 

### 2015-11-16 src/Main.hs
Refactor

@@ -44,7 +44,7 @@ atTag tag = deep (isElem >>> hasName tag)
 words :: ArrowXml cat => cat XmlTree (String, String)
 words = atTag "S" >>>
   proc x -> do
-    w       <- (atTag "W") -< x
+    w       <- atTag "W" -< x
     word    <- getText <<< getChildren -< w
     tagStr  <- getAttrValue "IG" -< w
     returnA -< (word, tagStr)

### 2015-11-17 proposal/proposal.tex
Update proposal

@@
 \usepackage{hyperref}
+
+\newcommand{\hmmURL}{https://hackage.haskell.org/package/hmm-0.2.1.1/docs/Data-HMM.html}
 
@@
 The most important advancement in the computational linguistics of Turkish is
-the NLP framework ``Zemberek'' by \citet{akin2007zemberek}. Although it started
+the NLP framework ``Zemberek'' by \citet{akin2007zemberek} that has been
+immensely useful for both research and applications of Turkish NLP. Although it started
 out as an NLP tool for the Turkish language, Zemberek moved on to become a
-a tool for processing all Turkic languages. Zemberek has been immensely useful
-for linguistic research on Turkish.
+a tool for processing all Turkic languages. Zemberek
 
@@
 
-Most of the work relating to the computational processing of Turkish deal with
-Turkish morphology since Turkish is a morphologically rich language analyzing
-the morphology seems to be the more salient problem e
+Most of the literature relating to the computational processing of Turkish
+pertain primarily to Turkish morphology---compared to the complexity of Turkish
+morphology, its syntax seems to be a less interesting problem.
 \section{Proposal}
@@
 I will be implementing a part-of-speech tagger for the Turkish language. As
-mention in section~\ref{sec:related_work} I will be mostly focusing on parts-of-speech
-of words and will not be accounting for morphology. There are problems with this
+mentioned in section~\ref{sec:related_work} I will be mostly focusing on parts-of-speech
+of words\footnote{I base my defition of a ``word'' on the orthographic space in
+  most of the cases. Some of the exceptions to this are Turkish reduplication
+  that have the POS ``Dup''} and will not be accounting for morphology. There are problems with this
 approach but this project aims to be a starting point for further research.
+
+For programming the POS tagger I will be using Haskell, a statically-typed and purely
+functional programming language. I will be using the
+\href{\hmmURL}{\texttt{Data.HMM}} package for
+Haskell. The data, that is probably the most important component of this
+project, comes from the METU-Sabanc\i treebank built by
+\citet{oflazer2003building}. I will be stripping the POS tag for each word from
+the corpus and will use them to train HMM's.
 

### 2015-11-17 proposal/proposal.tex
Update proposal

@@
 My main motivation for undertaking this project is to understand the main
-principles underlying POS-tagging; due to time and resource constraints, it is
+principles underlying POS-tagging. Due to time and resource constraints, it is
 not possible for me to build a state-of-the-art POS-tagger for
 Turkish. Nevertheless, I expect my POS-tagger
-to work with a decent accuracy that is enough to be a starting point for
+to work with a decent accuracy that is adequate to be a starting point for
 further development. Moreover, \citet{Korkut2015} began the development of a
-Turkish NLP framework named ``Guguk'' implemented in the programming language Haskell. I intend
-to incorporate this project to the Guguk project, hence contributing towards
+Turkish NLP framework named ``Guguk,'' implemented in the programming language Haskell. I intend
+to merge this project to Guguk, hence contributing towards
 the existence of a Turkish NLP-framework in Haskell.
@@
 the NLP framework ``Zemberek'' by \citet{akin2007zemberek} that has been
-immensely useful for both research and applications of Turkish NLP. Although it started
+immensely useful for both theory and applications of Turkish NLP. Although it started
 out as an NLP tool for the Turkish language, Zemberek moved on to become a
-a tool for processing all Turkic languages. Zemberek
+a tool for processing all Turkic languages.
 
@@
 
-I will be implementing a part-of-speech tagger for the Turkish language. As
-mentioned in section~\ref{sec:related_work} I will be mostly focusing on parts-of-speech
+I will implement a part-of-speech tagger for the Turkish language. As
+mentioned in section~\ref{sec:related_work}, I will mostly focus on parts-of-speech
 of words\footnote{I base my defition of a ``word'' on the orthographic space in
@@
   that have the POS ``Dup''} and will not be accounting for morphology. There are problems with this
-approach but this project aims to be a starting point for further research.
+approach, fixing which would go beyond the scope of this class.
 

### 2015-11-17 proposal/proposal.tex
Fix minor stuff

@@
 \label{sec:motivation}
-My main motivation for undertaking this project is to understand the main
+My main motivation for undertaking this project is to understand the
 principles underlying POS-tagging. Due to time and resource constraints, it is
@@
 The most important advancement in the computational linguistics of Turkish is
-the NLP framework ``Zemberek'' by \citet{akin2007zemberek} that has been
-immensely useful for both theory and applications of Turkish NLP. Although it started
+probably the NLP framework ``Zemberek'' by \citet{akin2007zemberek}, which has been
+immensely useful for both the theory and practice of Turkish NLP. Although it started
 out as an NLP tool for the Turkish language, Zemberek moved on to become a
@@
 
-Most of the literature relating to the computational processing of Turkish
-pertain primarily to Turkish morphology---compared to the complexity of Turkish
+In general, most of the literature relating to the computational processing of Turkish
+pertain primarily to morphology---compared to the complexity of Turkish
 morphology, its syntax seems to be a less interesting problem.
@@
   most of the cases. Some of the exceptions to this are Turkish reduplication
-  that have the POS ``Dup''} and will not be accounting for morphology. There are problems with this
-approach, fixing which would go beyond the scope of this class.
+  that have the POS ``Dup''} and will not be accounting for morphology, unlike
+the case with most of the serious research on Turkish NLP. There definitely are
+problems with this approach---but fixing these would go beyond the scope of
+this class
 
 For programming the POS tagger I will be using Haskell, a statically-typed and purely
-functional programming language. I will be using the
-\href{\hmmURL}{\texttt{Data.HMM}} package for
-Haskell. The data, that is probably the most important component of this
+functional programming language, along with the
+\href{\hmmURL}{\texttt{Data.HMM}} package.
+The data, that is probably the most important component of this
 project, comes from the METU-Sabanc\i treebank built by

### 2015-11-17 proposal/proposal.tex
Fix footnote

@@
 of words\footnote{I base my defition of a ``word'' on the orthographic space in
-  most of the cases. Some of the exceptions to this are Turkish reduplication
-  that have the POS ``Dup''} and will not be accounting for morphology, unlike
+  most of the cases. There are certain exceptions to this, such as the
+  reduplication which I denote with the POS tag ``Dup''} and will not be accounting for morphology, unlike
 the case with most of the serious research on Turkish NLP. There definitely are

### 2015-11-17 proposal/proposal.tex
Fix typo

--- 2015-11-17 proposal/proposal.tex
@@
 of words\footnote{I base my defition of a ``word'' on the orthographic space in
-  most of the cases. Some of the exceptions to this are Turkish reduplication
-  that have the POS ``Dup''} and will not be accounting for morphology, unlike
+  most of the cases. There are certain exceptions to this, such as
+  reduplication which I denote with the POS tag ``Dup''} and will not be accounting for morphology, unlike
 the case with most of the serious research on Turkish NLP. There definitely are

### 2015-11-17 proposal/proposal.tex
Fix things

@@
 The data, that is probably the most important component of this
-project, comes from the METU-Sabanc\i treebank built by
-\citet{oflazer2003building}.  I will be stripping the POS tag for each word from
-the corpus and will use them to train HMM's.
+project, comes from the METU-Sabanc{\i} treebank built by
+\citet{oflazer2003building}.  I am planning to strip the POS tag for each word from
+the treebank and use those to train the HMM
 

### 2015-11-17 proposal/proposal.tex
Run file through aspell

@@
 mentioned in section~\ref{sec:related_work}, I will mostly focus on parts-of-speech
-of words\footnote{I base my defition of a ``word'' on the orthographic space in
+of words\footnote{I base my definition of a ``word'' on the orthographic space in
   most of the cases. There are certain exceptions to this, such as

### 2015-11-17 src/Main.hs
Add ability to take an arbitrary size word-tag pair from the treebank

@@ -2,7 +2,7 @@
 
 module Main where
 
--- import Data.HMM
+import Data.HMM
 import System.IO hiding (putStr)
 import System.Directory
 -- import qualified Data.Vector.Storable as VS
@@ -52,10 +52,22 @@ words = atTag "S" >>>
 printTuple :: (String, String) -> IO ()
 printTuple (a, b)  = putStrLn (a ++ " " ++ (show $ extractPOS b))
 
+-- | Given number of files `numFiles` takes out all word-tag pairs
+-- | from each file indexed from 0 .. `numFiles`, and concatenates
+-- | them.
+-- getWords :: ArrowXml cat => FilePath -> cat XmlTree (String, String)
+getWords f = readDocument [withValidate no] f >>> words
+
 main :: IO ()
 main = do
   files <- getDirectoryContents "tb_uni"
-  let file = "tb_uni/" ++ files !! 3
-  ws <- runX (readDocument [withValidate no] file >>> words)
-  _  <- mapM printTuple ws
+  let file = "tb_uni/" ++ files !! 4
+      fileNames = drop 2 . take 40 $ map ("tb_uni/" ++) files
+  -- pairs <- runX $ readDocument [withValidate no] file >>> words
+  pairList <- mapM runX $ map getWords fileNames
+  print pairList
+  -- let words = map fst pairs
+      -- tags  = map (extractPOS . snd) pairs
+      -- myHMM = simpleHMM words tags
+  -- print myHMM
   return ()

### 2015-11-17 src/Main.hs
Trying to set up HMM

@@ -28,6 +28,9 @@ data POS = Noun
          | Punc
          deriving (Eq, Show, Read)
 
+allTags :: [POS]
+allTags = [Noun,Adj,Adv,Verb,Pron,Conj,Det,Postp,Ques,Interj,Num,Dup,Punc]
+
 extractPOS :: String -> POS
 extractPOS x = read $ takeWhile notPlusNorQuote . tail $ dropWhile notPlus x
   where notPlus = (/= '+')
@@ -62,12 +65,13 @@ main :: IO ()
 main = do
   files <- getDirectoryContents "tb_uni"
   let file = "tb_uni/" ++ files !! 4
-      fileNames = drop 2 . take 40 $ map ("tb_uni/" ++) files
+      fileNames = drop 2 . take 100 $ map ("tb_uni/" ++) files
   -- pairs <- runX $ readDocument [withValidate no] file >>> words
   pairList <- mapM runX $ map getWords fileNames
-  print pairList
-  -- let words = map fst pairs
-      -- tags  = map (extractPOS . snd) pairs
-      -- myHMM = simpleHMM words tags
-  -- print myHMM
+  let pairs      = foldr (++) [] pairList
+      words      = map fst pairs
+      tags       = map (extractPOS . snd) pairs
+      myHMM      = simpleHMM allTags []
+      -- trainedHmm = baumWelch myHMM words 10
+  _ <- print trainedHmm
   return ()

### 2015-11-18 src/Main.hs
Enumerate parts of speech

@@ -26,7 +26,13 @@ data POS = Noun
          | Num
          | Dup
          | Punc
-         deriving (Eq, Show, Read)
+         deriving (Eq, Show, Read, Ord, Enum)
+
+posToInt :: POS -> Int
+posToInt p = p `elemIndex` [Noun .. Punc]
+
+intToPos :: Int -> POS
+intToPos i = [Noun .. Punc] !! i
 
 allTags :: [POS]
 allTags = [Noun,Adj,Adv,Verb,Pron,Conj,Det,Postp,Ques,Interj,Num,Dup,Punc]
@@ -73,5 +79,7 @@ main = do
       tags       = map (extractPOS . snd) pairs
       myHMM      = simpleHMM allTags []
       -- trainedHmm = baumWelch myHMM words 10
-  _ <- print trainedHmm
+  print pairList
+  -- print $ zip words tags
+  -- print $ [Noun .. Punc]
   return ()

### 2015-11-18 src/Main.hs
Use Enum typeclass functions

@@ -7,6 +7,8 @@ import System.IO hiding (putStr)
 import System.Directory
 -- import qualified Data.Vector.Storable as VS
 import Text.XML.HXT.Core
+import Data.Array
+import Data.List (elemIndex)
 import Prelude hiding (words, putStr)
 
 type TaggedWord = (String, POS)
@@ -28,12 +30,6 @@ data POS = Noun
          | Punc
          deriving (Eq, Show, Read, Ord, Enum)
 
-posToInt :: POS -> Int
-posToInt p = p `elemIndex` [Noun .. Punc]
-
-intToPos :: Int -> POS
-intToPos i = [Noun .. Punc] !! i
-
 allTags :: [POS]
 allTags = [Noun,Adj,Adv,Verb,Pron,Conj,Det,Postp,Ques,Interj,Num,Dup,Punc]
 
@@ -45,11 +41,6 @@ extractPOS x = read $ takeWhile notPlusNorQuote . tail $ dropWhile notPlus x
 atTag :: ArrowXml a => String -> a XmlTree XmlTree
 atTag tag = deep (isElem >>> hasName tag)
 
--- | For all children of <S> (i.e., the words) return a tuple containing
--- | the word and the string containing part-of-speech and morpheme
--- | glossing of the word. We are primarily interested in the part-of-speech
--- | contained by this `tagStr` so we will eventually extract it out using
--- | `extractPOS`.
 words :: ArrowXml cat => cat XmlTree (String, String)
 words = atTag "S" >>>
   proc x -> do
@@ -77,9 +68,13 @@ main = do
   let pairs      = foldr (++) [] pairList
       words      = map fst pairs
       tags       = map (extractPOS . snd) pairs
-      myHMM      = simpleHMM allTags []
-      -- trainedHmm = baumWelch myHMM words 10
-  print pairList
+      -- Now let us "index" each of the words with a
+      -- tag index i.e. (fromEnum t) where t ∈ POS.
+      pairs'     = zip (map fromEnum tags) words
+      pairsArr   = array (0, (length pairs')-1) pairs'
+      myHMM      = simpleHMM [0..12] words
+      trainedHmm = baumWelch myHMM pairsArr 10
+  print trainedHmm
   -- print $ zip words tags
   -- print $ [Noun .. Punc]
   return ()

### 2015-11-18 src/Main.hs
Remove redundant stuff

@@ -30,9 +30,6 @@ data POS = Noun
          | Punc
          deriving (Eq, Show, Read, Ord, Enum)
 
-allTags :: [POS]
-allTags = [Noun,Adj,Adv,Verb,Pron,Conj,Det,Postp,Ques,Interj,Num,Dup,Punc]
-
 extractPOS :: String -> POS
 extractPOS x = read $ takeWhile notPlusNorQuote . tail $ dropWhile notPlus x
   where notPlus = (/= '+')

### 2015-11-18 src/Main.hs
Type signature for top level

@@ -52,7 +52,7 @@ printTuple (a, b)  = putStrLn (a ++ " " ++ (show $ extractPOS b))
 -- | Given number of files `numFiles` takes out all word-tag pairs
 -- | from each file indexed from 0 .. `numFiles`, and concatenates
 -- | them.
--- getWords :: ArrowXml cat => FilePath -> cat XmlTree (String, String)
+getWords :: FilePath -> IOSLA (XIOState s) a (String, String)
 getWords f = readDocument [withValidate no] f >>> words
 
 main :: IO ()
@@ -60,7 +60,6 @@ main = do
   files <- getDirectoryContents "tb_uni"
   let file = "tb_uni/" ++ files !! 4
       fileNames = drop 2 . take 100 $ map ("tb_uni/" ++) files
-  -- pairs <- runX $ readDocument [withValidate no] file >>> words
   pairList <- mapM runX $ map getWords fileNames
   let pairs      = foldr (++) [] pairList
       words      = map fst pairs

### 2015-11-18 src/Main.hs
Separate modules

@@ -1,59 +1,15 @@
-{-# LANGUAGE Arrows, NoMonomorphismRestriction #-}
 
 module Main where
 
 import Data.HMM
 import System.IO hiding (putStr)
 import System.Directory
--- import qualified Data.Vector.Storable as VS
-import Text.XML.HXT.Core
 import Data.Array
+import Text.XML.HXT.Core
 import Data.List (elemIndex)
 import Prelude hiding (words, putStr)
+import Parse
 
-type TaggedWord = (String, POS)
-
--- | These are the possible parts-of-speech described in the
--- | METU-Sabanci treebank paper.
-data POS = Noun
-         | Adj
-         | Adv
-         | Verb
-         | Pron
-         | Conj
-         | Det
-         | Postp
-         | Ques
-         | Interj
-         | Num
-         | Dup
-         | Punc
-         deriving (Eq, Show, Read, Ord, Enum)
-
-extractPOS :: String -> POS
-extractPOS x = read $ takeWhile notPlusNorQuote . tail $ dropWhile notPlus x
-  where notPlus = (/= '+')
-        notPlusNorQuote = \n -> notPlus n && n /= '"'
-
-atTag :: ArrowXml a => String -> a XmlTree XmlTree
-atTag tag = deep (isElem >>> hasName tag)
-
-words :: ArrowXml cat => cat XmlTree (String, String)
-words = atTag "S" >>>
-  proc x -> do
-    w       <- atTag "W" -< x
-    word    <- getText <<< getChildren -< w
-    tagStr  <- getAttrValue "IG" -< w
-    returnA -< (word, tagStr)
-
-printTuple :: (String, String) -> IO ()
-printTuple (a, b)  = putStrLn (a ++ " " ++ (show $ extractPOS b))
-
--- | Given number of files `numFiles` takes out all word-tag pairs
--- | from each file indexed from 0 .. `numFiles`, and concatenates
--- | them.
-getWords :: FilePath -> IOSLA (XIOState s) a (String, String)
-getWords f = readDocument [withValidate no] f >>> words
 
 main :: IO ()
 main = do
@@ -62,15 +18,18 @@ main = do
       fileNames = drop 2 . take 100 $ map ("tb_uni/" ++) files
   pairList <- mapM runX $ map getWords fileNames
   let pairs      = foldr (++) [] pairList
-      words      = map fst pairs
+      ws         = map fst pairs
       tags       = map (extractPOS . snd) pairs
-      -- Now let us "index" each of the words with a
[…]

### 2015-11-18 src/Parse.hs

@@ -1,0 +1,47 @@
+{-# LANGUAGE Arrows, NoMonomorphismRestriction #-}
+
+module Parse where
+
+import Prelude hiding (words)
+import Text.XML.HXT.Core
+-- | These are the possible parts-of-speech described in the
+-- | METU-Sabanci treebank paper.
+data POS = Noun
+         | Adj
+         | Adv
+         | Verb
+         | Pron
+         | Conj
+         | Det
+         | Postp
+         | Ques
+         | Interj
+         | Num
+         | Dup
+         | Punc
+         deriving (Eq, Show, Read, Ord, Enum)
+
+extractPOS :: String -> POS
+extractPOS x = read $ takeWhile notPlusNorQuote . tail $ dropWhile notPlus x
+  where notPlus = (/= '+')
+        notPlusNorQuote = \n -> notPlus n && n /= '"'
+
+atTag :: ArrowXml a => String -> a XmlTree XmlTree
+atTag tag = deep (isElem >>> hasName tag)
+
+words :: ArrowXml cat => cat XmlTree (String, String)
+words = atTag "S" >>>
+  proc x -> do
+    w       <- atTag "W" -< x
+    word    <- getText <<< getChildren -< w
+    tagStr  <- getAttrValue "IG" -< w
+    returnA -< (word, tagStr)
+
+printTuple :: (String, String) -> IO ()
+printTuple (a, b)  = putStrLn (a ++ " " ++ (show $ extractPOS b))
+
+-- | Given number of files `numFiles` takes out all word-tag pairs
+-- | from each file indexed from 0 .. `numFiles`, and concatenates
+-- | them.
+getWords :: FilePath -> IOSLA (XIOState s) a (String, String)
+getWords f = readDocument [withValidate no] f >>> words

### 2015-11-18 src/Main.hs
Avoid warnings

@@ -6,16 +6,13 @@ import System.IO hiding (putStr)
 import System.Directory
 import Data.Array
 import Text.XML.HXT.Core
-import Data.List (elemIndex)
 import Prelude hiding (words, putStr)
 import Parse
-
 
 main :: IO ()
 main = do
   files <- getDirectoryContents "tb_uni"
-  let file = "tb_uni/" ++ files !! 4
-      fileNames = drop 2 . take 100 $ map ("tb_uni/" ++) files
+  let fileNames = drop 2 . take 100 $ map ("tb_uni/" ++) files
   pairList <- mapM runX $ map getWords fileNames
   let pairs      = foldr (++) [] pairList
       ws         = map fst pairs
@@ -25,11 +22,7 @@ main = do
       pairs'     = zip (map fromEnum tags) ws
       pairsArr   = array (0, 1386) $ zip [0..1386] ws
       myHMM      = simpleHMM [0..12] ws
-      trainedHMM = baumWelch myHMM pairsArr 30
-      sampleSent = array (0, 5) $ zip [0..5] [ "yılının"
-                                             , "ilk"
-                                             , "ayına"
-                                             , "iliskin"]
-  saveHMM "model.hmm" trainedHMM
+      trainedHMM = baumWelch myHMM pairsArr 2
+  saveHMM "foo.hmm" trainedHMM
   -- _ <- mapM putStrLn [pairsArr ! x | x <- [0..15]]
   return ()

### 2015-11-18 tag/Main.hs
Multiple executables

@@ -1,0 +1,14 @@
+module Main where
+import Data.HMM (loadHMM, viterbi, HMM(..))
+import System.IO
+import Data.Array
+
+main :: IO ()
+main = do
+  -- trainedHMM <- (loadHMM "model.hmm") :: IO (HMM Int String)
+  putStr "> "
+  hFlush stdout
+  input      <- getLine
+  let ws      = words input
+      wsArray = array (0, )
+  return ()

### 2015-11-18 tag/Main.hs
Receive input

@@ -1,3 +1,5 @@
+{-# LANGUAGE BangPatterns #-}
+
 module Main where
 import Data.HMM (loadHMM, viterbi, HMM(..))
 import System.IO
@@ -8,7 +10,8 @@ main = do
   -- trainedHMM <- (loadHMM "model.hmm") :: IO (HMM Int String)
   putStr "> "
   hFlush stdout
-  input      <- getLine
-  let ws      = words input
-      wsArray = array (0, )
+  input <- getLine
+  let ws       = words input
+      !wsArray = listArray (0, (length ws)-1) ws
+  print wsArray
   return ()

### 2015-11-19 tag/Main.hs
Change IO

@@ -7,11 +7,12 @@ import Data.Array
 
 main :: IO ()
 main = do
-  -- trainedHMM <- (loadHMM "model.hmm") :: IO (HMM Int String)
+  trainedHMM <- (loadHMM "model.hmm") :: IO (HMM Int String)
   putStr "> "
   hFlush stdout
   input <- getLine
   let ws       = words input
       !wsArray = listArray (0, (length ws)-1) ws
-  print wsArray
+      !out      = viterbi trainedHMM wsArray
+  print $ head out
   return ()

### 2015-11-19 paper.tex

\documentclass{article}
\usepackage{natbib}
\usepackage{todonotes}

\title{\bf Final Project Proposal}
\author{Your Name}
\date{Due date: 8 December 2015 at 11:59 PM Central time}

\begin{document}
\maketitle

% Lines that begin with % are comments.

\section{Motivation}
\label{sec:motivation}

% Delete this todo
\todo[inline]{
Write an introductory section that motivates a particular problem you wish to solve. This section should (at a minimum) fill the remainder of this page.
}

\section{Related Work}

% Delete this todo
\todo[inline]{%
Briefly survey the most salient prior work that relates to the problem you wish to solve. This section should cite relevant sources. This section should be at half a page in length.
}

\section{Proposed Work}

% Delete this todo
\todo[inline]{This section should contain a detailed description of the work you propose to do, but it should not contain any experimental results. This section should take at least a page and a half.}

\section{Results}

% Delete this todo
\todo[inline]{This section should describe the your results. In other words, after you do what you proposed in section 3, what results did you come up with? This section should be at least one page in length.
}

\section{Conclusion}

% Delete this todo
\todo[inline]{Write a concluding summary. Also indicate what future work may be warranted. This section should be at least half a page in length.}

\bibliographystyle{apalike}
\bibliography{your_bibliography}

\end{document}

### 2015-11-21 create/Main.hs
Frequency map and some UnicodeSyntax

@@ -1,28 +1,48 @@
-
+{-# LANGUAGE UnicodeSyntax #-}
 module Main where
 
-import Data.HMM
+-- import Data.HMM
 import System.IO hiding (putStr)
 import System.Directory
-import Data.Array
+-- import Data.Array
 import Text.XML.HXT.Core
 import Prelude hiding (words, putStr)
+import qualified Data.Map as M
 import Parse
 
-main :: IO ()
+type ℤ = Int
+type ℚ = Double
+
+freqMap ∷ Ord a ⇒ [a] → M.Map a ℤ
+freqMap unigrams = populate M.empty unigrams
+  where populate ∷ Ord a ⇒ M.Map a ℤ → [a] → M.Map a ℤ
+        populate m [] = m
+        populate m (x:xs) = let freq = (M.findWithDefault 0 x m) :: ℤ
+                            in populate (M.insert x (freq + 1) m) xs
+
+
+main ∷ IO ()
 main = do
-  files <- getDirectoryContents "tb_uni"
+  files ← getDirectoryContents "tb_uni"
   let fileNames = drop 2 . take 100 $ map ("tb_uni/" ++) files
-  pairList <- mapM runX $ map getWords fileNames
-  let pairs      = foldr (++) [] pairList
-      ws         = map fst pairs
-      tags       = map (extractPOS . snd) pairs
-                   -- Now let us "index" each of the words with a
-                   -- tag index i.e. (fromEnum t) where t ∈ POS.
-      pairs'     = zip (map fromEnum tags) ws
-      pairsArr   = array (0, 1386) $ zip [0..1386] ws
-      myHMM      = simpleHMM [0..12] ws
-      trainedHMM = baumWelch myHMM pairsArr 2
-  saveHMM "foo.hmm" trainedHMM
-  -- _ <- mapM putStrLn [pairsArr ! x | x <- [0..15]]
+  pairList ← mapM runX $ map getWords fileNames
+      -- First we need to get POS types from the textual representation in the
+      -- corpus. the `pairList` that we have right now is a list of
+      -- list of tuples. We call `extractPOS` on the `snd` element of
+      -- each tuple.
+  let tagsList = map (\xs → map (extractPOS . snd) xs) pairList
+      -- Now let us append a beginning of sentence POS to each of the lists
+      -- in tagsList and denote this with tagsList'.
+      tagsList'      ∷ [[POS]]
[…]

### 2015-11-21 create/Parse.hs

@@ -6,7 +6,8 @@ import Prelude hiding (words)
 import Text.XML.HXT.Core
 -- | These are the possible parts-of-speech described in the
 -- | METU-Sabanci treebank paper.
-data POS = Noun
+data POS = Start
+         | Noun
          | Adj
          | Adv
          | Verb

### 2015-11-21 create/Main.hs
Tag unigram freqs

@@ -35,14 +35,11 @@ main = do
       -- in tagsList and denote this with tagsList'.
       tagsList'      ∷ [[POS]]
       tagsList'      = map ((:) Start) tagsList
-      tagsList''     = foldr (++) [] tagsList'
-      tagBigrams     = [(tagsList'' !! i, tagsList'' !! (i+1))
-                         | i ← [0..length tagsList'' - 2]]
+      -- If we concat all the lists, we're left with just list of tags.
+      tags           = foldr (++) [] tagsList'
+      tagBigrams     = [(tags !! i, tags !! (i+1))
+                         | i ← [0..length tags - 2]]
       tagBigramFreqs = freqMap tagBigrams
-      -- Make a transformation matrix out of this list.
-      -- transMatrix ∷ M.Map (POS, POS) Double
-      -- pairs'     = zip (map fromEnum tags) ws
-      -- pairsArr   = array (0, 1386) $ zip [0..1386] ws
+      tagFreqs       = freqMap tags
   print $ tagBigramFreqs M.! (Noun, Verb)
-  -- print $ tagBigramFreqs M.! (Det, Det)
-  return ()
+  print $ tagFreqs M.! Noun

### 2015-11-21 create/Main.hs
Add TODO stub for probability

@@ -17,9 +17,13 @@ freqMap ∷ Ord a ⇒ [a] → M.Map a ℤ
 freqMap unigrams = populate M.empty unigrams
   where populate ∷ Ord a ⇒ M.Map a ℤ → [a] → M.Map a ℤ
         populate m [] = m
-        populate m (x:xs) = let freq = (M.findWithDefault 0 x m) :: ℤ
+        populate m (x:xs) = let freq = (M.findWithDefault 0 x m) ∷ ℤ
                             in populate (M.insert x (freq + 1) m) xs
 
+-- | Probability of a, given b.
+-- | TODO: Implement
+probability :: POS → POS → ℚ
+probability a b = undefined
 
 main ∷ IO ()
 main = do

### 2015-11-22 create/Main.hs
Add function for computing POS probability

@@ -10,8 +10,8 @@ import Prelude hiding (words, putStr)
 import qualified Data.Map as M
 import Parse
 
-type ℤ = Int
-type ℚ = Double
+type ℤ = Int    -- The integer type `Int` will be denoted by ℤ.
+type ℚ = Double -- The double type will be denoted by ℚ.
 
 freqMap ∷ Ord a ⇒ [a] → M.Map a ℤ
 freqMap unigrams = populate M.empty unigrams
@@ -20,10 +20,10 @@ freqMap unigrams = populate M.empty unigrams
         populate m (x:xs) = let freq = (M.findWithDefault 0 x m) ∷ ℤ
                             in populate (M.insert x (freq + 1) m) xs
 
--- | Probability of a, given b.
--- | TODO: Implement
-probability :: POS → POS → ℚ
-probability a b = undefined
+posProbability ∷ POS → POS → M.Map POS ℤ → M.Map (POS, POS) ℤ → ℚ
+posProbability a b m₁ m₂ = fromIntegral bigramFreq / fromIntegral unigramFreq
+  where unigramFreq = M.findWithDefault 0 a m₁
+        bigramFreq  = M.findWithDefault 1 (a, b) m₂
 
 main ∷ IO ()
 main = do
@@ -45,5 +45,4 @@ main = do
                          | i ← [0..length tags - 2]]
       tagBigramFreqs = freqMap tagBigrams
       tagFreqs       = freqMap tags
-  print $ tagBigramFreqs M.! (Noun, Verb)
-  print $ tagFreqs M.! Noun
+  print tagBigramFreqs

### 2015-11-22 create/Main.hs
Change some variable names for clarification

@@ -21,9 +21,9 @@ freqMap unigrams = populate M.empty unigrams
                             in populate (M.insert x (freq + 1) m) xs
 
 posProbability ∷ POS → POS → M.Map POS ℤ → M.Map (POS, POS) ℤ → ℚ
-posProbability a b m₁ m₂ = fromIntegral bigramFreq / fromIntegral unigramFreq
-  where unigramFreq = M.findWithDefault 0 a m₁
-        bigramFreq  = M.findWithDefault 1 (a, b) m₂
+posProbability t tₙₑₓₜ m₁ m₂ = fromIntegral bigramCount / fromIntegral tCount
+  where tCount      = M.findWithDefault 0 t m₁
+        bigramCount = M.findWithDefault 1 (t, tₙₑₓₜ) m₂
 
 main ∷ IO ()
 main = do

### 2015-11-22 create/Main.hs
Push some bigram frequencies

@@ -20,10 +20,10 @@ freqMap unigrams = populate M.empty unigrams
         populate m (x:xs) = let freq = (M.findWithDefault 0 x m) ∷ ℤ
                             in populate (M.insert x (freq + 1) m) xs
 
-posProbability ∷ POS → POS → M.Map POS ℤ → M.Map (POS, POS) ℤ → ℚ
-posProbability t tₙₑₓₜ m₁ m₂ = fromIntegral bigramCount / fromIntegral tCount
-  where tCount      = M.findWithDefault 0 t m₁
-        bigramCount = M.findWithDefault 1 (t, tₙₑₓₜ) m₂
+bigramProbability ∷ (POS, POS) → M.Map POS ℤ → M.Map (POS, POS) ℤ → ℚ
+bigramProbability (t, tₙₑₓₜ) m₁ m₂ = bigramCount / tCount
+  where tCount      = fromIntegral (m₁ M.! t) ∷ ℚ
+        bigramCount = fromIntegral (M.findWithDefault 0 (t, tₙₑₓₜ) m₂) ∷ ℚ
 
 main ∷ IO ()
 main = do
@@ -34,15 +34,18 @@ main = do
       -- corpus. the `pairList` that we have right now is a list of
       -- list of tuples. We call `extractPOS` on the `snd` element of
       -- each tuple.
-  let tagsList = map (\xs → map (extractPOS . snd) xs) pairList
+  let tagsList = map (map (extractPOS . snd)) pairList
       -- Now let us append a beginning of sentence POS to each of the lists
       -- in tagsList and denote this with tagsList'.
       tagsList'      ∷ [[POS]]
       tagsList'      = map ((:) Start) tagsList
       -- If we concat all the lists, we're left with just list of tags.
-      tags           = foldr (++) [] tagsList'
+      tags           = concat tagsList'
       tagBigrams     = [(tags !! i, tags !! (i+1))
                          | i ← [0..length tags - 2]]
       tagBigramFreqs = freqMap tagBigrams
       tagFreqs       = freqMap tags
-  print tagBigramFreqs
+  print $ bigramProbability (Start, Noun) tagFreqs tagBigramFreqs
+  print $ bigramProbability (Start, Verb) tagFreqs tagBigramFreqs
+  print $ bigramProbability (Start, Pron) tagFreqs tagBigramFreqs
+  print $ bigramProbability (Noun,  Verb) tagFreqs tagBigramFreqs

### 2015-12-01 create/Main.hs
Write model to a file

@@ -2,7 +2,7 @@
 
 module Main where
 
-import Data.HMM (viterbi, HMM(..), saveHMM)
+import Data.HMM (viterbi, HMM(..))
 import System.IO
 import System.Directory
 import Data.Array (listArray, Array)
@@ -84,5 +84,6 @@ main = do
                                            , "beni"
                                            , "hatırlasın"
                                            , "."]
+  writeFile "model.hmm" (show newHMM)
   print $ viterbi newHMM sampleSentence₅
   return ()

### 2015-12-04 create/Main.hs
Use LogFloats

@@ -2,18 +2,18 @@
 
 module Main where
 
-import Data.HMM (viterbi, HMM(..))
-import System.IO
-import System.Directory
-import Data.Array (listArray, Array)
-import Text.XML.HXT.Core
-import Prelude hiding (words, putStr)
-import qualified Data.Map as M
-import Data.Number.LogFloat
-import Parse
+import           Data.Array           (Array, listArray)
+import           Data.HMM             (HMM (..), viterbi)
+import qualified Data.Map             as M
+import           Data.Number.LogFloat
+import           Parse
+import           Prelude              hiding (putStr, words)
+import           System.Directory
+import           Text.XML.HXT.Core
 
-type ℤ = Int    -- The integer type `Int` will be denoted by ℤ.
-type ℚ = Double -- The double type will be denoted by ℚ.
+type ℤ = Int      -- The integer type `Int` will be denoted by ℤ.
+type ℚ = Double   -- The double type will be denoted by ℚ.
+type 𝓛 = LogFloat -- Log-domain numbers to prevent underflow.
 
 freqMap ∷ Ord a ⇒ [a] → M.Map a ℤ
 freqMap unigrams = populate M.empty unigrams
@@ -22,10 +22,13 @@ freqMap unigrams = populate M.empty unigrams
         populate m (x:xs) = let freq = (M.findWithDefault 0 x m) ∷ ℤ
                             in populate (M.insert x (freq + 1) m) xs
 
-probability ∷ (Ord a, Ord b) ⇒ (a, b) → M.Map a ℤ → M.Map (a, b) ℤ → ℚ
+--  | Takes in a tuple (x, y) (i.e., a bigram), and takes in two frequency maps
+--    c₁ and c₂, c₁ for counting all the occurrences of x and the other for
+--    counting the number of bigrams.
+probability ∷ (Ord a, Ord b) ⇒ (a, b) → M.Map a ℤ → M.Map (a, b) ℤ → 𝓛
 probability (x, y) c₁ c₂ = if xCount == 0 || yCount == 0
                            then 0.0001
-                           else yCount / xCount
+                           else (logFloat yCount) / (logFloat xCount)
   where yCount = fromIntegral (M.findWithDefault 0 (x, y) c₂) ∷ ℚ
         xCount = fromIntegral (M.findWithDefault 0 x c₁) ∷ ℚ
 
@@ -45,8 +48,8 @@ main = do
[…]

### 2015-12-04 create/Parse.hs

@@ -1,6 +1,6 @@
-{-# LANGUAGE Arrows #-}
+{-# LANGUAGE Arrows                    #-}
 {-# LANGUAGE NoMonomorphismRestriction #-}
-{-# LANGUAGE UnicodeSyntax #-}
+{-# LANGUAGE UnicodeSyntax             #-}
 
 module Parse where
 
@@ -9,8 +9,7 @@ import Data.Char (isSpace, toLower)
 import Text.XML.HXT.Core
 -- | These are the possible parts-of-speech described in the
 -- | METU-Sabanci treebank paper.
-data POS = Start
-         | Noun
+data POS = Noun
          | Adj
          | Adv
          | Verb

### 2015-12-04 create/Main.hs
Avoid constraint defaulting warning

@@ -15,6 +15,41 @@ type ℤ = Int      -- The integer type `Int` will be denoted by ℤ.
 type ℚ = Double   -- The double type will be denoted by ℚ.
 type 𝓛 = LogFloat -- Log-domain numbers to prevent underflow.
 
+sampleSentence₁ ∷ Array Int String
+sampleSentence₁ = listArray (0, 4) [ "gözleri"
+                                    , "kor"
+                                    , "gibi"
+                                    , "yanıyordu"
+                                    , "."]
+
+sampleSentence₂ ∷ Array Int String
+sampleSentence₂ = listArray (0, 3) [ "adam"
+                                    , "yine"
+                                    , "geldi"
+                                    , "."
+                                    ]
+
+sampleSentence₃ ∷ Array Int String
+sampleSentence₃ = listArray (0, 4) [ "güzel"
+                                    , "kız"
+                                    , "mutlu"
+                                    , "gözüküyordu"
+                                    , "."]
+
+sampleSentence₄ ∷ Array Int String
+sampleSentence₄   = listArray (0, 5) [ "renksiz"
+                                    , "yeşil"
+                                    , "fikirler"
+                                    , "sessizce"
+                                    , "uyuyor"
+                                    , "."]
+
+sampleSentence₅ ∷ Array Int String
+sampleSentence₅   = listArray (0, 3) [ "dostlar"
+                                    , "beni"
+                                    , "hatırlasın"
+                                    , "."]
+
 freqMap ∷ Ord a ⇒ [a] → M.Map a ℤ
 freqMap unigrams = populate M.empty unigrams
   where populate ∷ Ord a ⇒ M.Map a ℤ → [a] → M.Map a ℤ
@@ -61,35 +96,9 @@ main = do
                              , initProbs   = initProbFn
                              , transMatrix = transFn
                              , outMatrix   = outFn}
-      sampleSentence   ∷ Array Int String
-      sampleSentence   = listArray (0, 4) [ "gözleri"
-                                          , "kor"
-                                          , "gibi"
-                                          , "yanıyordu"
-                                          , "."]
-      sampleSentence₂   = listArray (0, 3) [ "adam"
-                                           , "yine"
-                                           , "geldi"
-                                           , "."
-                                           ]
-      sampleSentence₃   = listArray (0, 4) [ "güzel"
-                                           , "kız"
-                                           , "mutlu"
-                                           , "gözüküyordu"
-                                           , "."]
[…]

### 2015-12-04 create/Main.hs
Pretty printing for tagged sentences

@@ -5,10 +5,12 @@ module Main where
 import           Data.Array           (Array, listArray)
 import           Data.HMM             (HMM (..), viterbi)
 import qualified Data.Map             as M
+import           Data.List            (intercalate)
 import           Data.Number.LogFloat
 import           Parse
-import           Prelude              hiding (putStr, words)
+import           Prelude              hiding (putStr, words, lines)
 import           System.Directory
+import           System.IO (putStr)
 import           Text.XML.HXT.Core
 
 type ℤ = Int      -- The integer type `Int` will be denoted by ℤ.
@@ -49,6 +51,12 @@ sampleSentence₅   = listArray (0, 3) [ "dostlar"
                                     , "beni"
                                     , "hatırlasın"
                                     , "."]
+
+printTaggedSent ∷ Array Int String → [POS] → IO ()
+printTaggedSent ws ps = let lines = (\n → replicate n '-') <$> (length <$> ws)
+                        in do putStrLn $ foldr (++) " "   $ (++ " ") <$> ws
+                              putStrLn $ foldr (++) " "   $ (++ " ") <$> lines
+                              putStrLn $ intercalate "  " $  map show ps
 
 freqMap ∷ Ord a ⇒ [a] → M.Map a ℤ
 freqMap unigrams = populate M.empty unigrams
@@ -98,8 +106,12 @@ main = do
                              , outMatrix   = outFn}
   writeFile "model.hmm" (show newHMM)
   putStrLn "Creating the model..."
-  print $ viterbi newHMM sampleSentence₁
-  print $ viterbi newHMM sampleSentence₂
-  print $ viterbi newHMM sampleSentence₃
-  print $ viterbi newHMM sampleSentence₄
-  print $ viterbi newHMM sampleSentence₅
+  printTaggedSent sampleSentence₁ $ viterbi newHMM sampleSentence₁
+  putStr "\n"
+  printTaggedSent sampleSentence₂ $ viterbi newHMM sampleSentence₂
+  putStr "\n"
+  printTaggedSent sampleSentence₃ $ viterbi newHMM sampleSentence₃
+  putStr "\n"
+  printTaggedSent sampleSentence₄ $ viterbi newHMM sampleSentence₄
+  putStr "\n"
+  printTaggedSent sampleSentence₅ $ viterbi newHMM sampleSentence₅

### 2015-12-04 create/Main.hs
Use minimum double for the case where a word is not found

@@ -17,6 +17,11 @@ type ℤ = Int      -- The integer type `Int` will be denoted by ℤ.
 type ℚ = Double   -- The double type will be denoted by ℚ.
 type 𝓛 = LogFloat -- Log-domain numbers to prevent underflow.
 
+ε ∷ Double
+ε = encodeFloat 1 $ fst  r - ds
+  where r  = floatRange (0.1 :: Double)
+        ds = floatDigits (0.1 :: Double)
+
 sampleSentence₁ ∷ Array Int String
 sampleSentence₁ = listArray (0, 4) [ "gözleri"
                                     , "kor"
@@ -70,7 +75,7 @@ freqMap unigrams = populate M.empty unigrams
 --    counting the number of bigrams.
 probability ∷ (Ord a, Ord b) ⇒ (a, b) → M.Map a ℤ → M.Map (a, b) ℤ → 𝓛
 probability (x, y) c₁ c₂ = if xCount == 0 || yCount == 0
-                           then 0.0001
+                           then logFloat ε
                            else (logFloat yCount) / (logFloat xCount)
   where yCount = fromIntegral (M.findWithDefault 0 (x, y) c₂) ∷ ℚ
         xCount = fromIntegral (M.findWithDefault 0 x c₁) ∷ ℚ

### 2015-12-04 create/Main.hs
Generate table

@@ -10,7 +10,7 @@ import           Data.Number.LogFloat
 import           Parse
 import           Prelude              hiding (putStr, words, lines)
 import           System.Directory
-import           System.IO (putStr)
+-- import           System.IO (putStr)
 import           Text.XML.HXT.Core
 
 type ℤ = Int      -- The integer type `Int` will be denoted by ℤ.
@@ -22,46 +22,95 @@ type 𝓛 = LogFloat -- Log-domain numbers to prevent underflow.
   where r  = floatRange (0.1 :: Double)
         ds = floatDigits (0.1 :: Double)
 
-sampleSentence₁ ∷ Array Int String
-sampleSentence₁ = listArray (0, 4) [ "gözleri"
-                                    , "kor"
-                                    , "gibi"
-                                    , "yanıyordu"
-                                    , "."]
+sample₁ ∷ Array Int String
+sample₁ = listArray (0, 4) [ "gözleri"
+                           , "kor"
+                           , "gibi"
+                           , "yanıyordu"
+                           , "."]
 
-sampleSentence₂ ∷ Array Int String
-sampleSentence₂ = listArray (0, 3) [ "adam"
-                                    , "yine"
-                                    , "geldi"
-                                    , "."
-                                    ]
+sample₂ ∷ Array Int String
+sample₂ = listArray (0, 3) [ "adam"
+                           , "yine"
+                           , "geldi"
+                           , "."
+                           ]
 
-sampleSentence₃ ∷ Array Int String
-sampleSentence₃ = listArray (0, 4) [ "güzel"
-                                    , "kız"
-                                    , "mutlu"
-                                    , "gözüküyordu"
-                                    , "."]
+sample₃ ∷ Array Int String
+sample₃ = listArray (0, 4) [ "güzel"
+                           , "kız"
+                           , "mutlu"
+                           , "gözüküyordu"
+                           , "."]
 
-sampleSentence₄ ∷ Array Int String
-sampleSentence₄   = listArray (0, 5) [ "renksiz"
-                                    , "yeşil"
-                                    , "fikirler"
-                                    , "sessizce"
-                                    , "uyuyor"
-                                    , "."]
+sample₄ ∷ Array Int String
+sample₄   = listArray (0, 5) [ "renksiz"
+                             , "yeşil"
+                             , "fikirler"
[…]

### 2015-12-04 paper/paper.tex
Update paper

--- 2015-11-17 proposal/proposal.tex
@@
 \usepackage{natbib}
+\usepackage{todonotes}
 \usepackage{hyperref}
@@
 \title{\bf An HMM-based part-of-speech tagger for Turkish}
-\author{Ayberk Tosun\\\texttt{tosun2@illinois.edu}}
+\author{Ayberk Tosun}
 \date{}
@@
 
+\section{Results}
+
+The POS tagger I've built as proposed can successfully tag words with a
+reasonable accuracy.
+
+\subsection{Limitations}
+There is a fundamental limitation to the POS tagger I built: Turkish morphology
+is essential to the overall expressivity of the Turkish language. In English we
+entertain the notion of ``syntactic correctness'', which we define as words of a
+sentence having some sort of correct order. A speaker of English that has
+synactic competence in the language \citep{bachman1990} may know that a sentence
+is well-formed, relying on his knowledge of syntactic rules. However, for
+\textit{words}, knowledge of morphological rules is not adequate to conclude
+that a word is valid; they have to rely on their lexical knowledge to verify
+that the morphological form they derived/inflected is a valid one. Consider the
+word ``satisfiableness'' which is not a valid word, derived with valid suffixes.
+The correct form of this concept is ``satisfiability''. In Turkish, the concept
+of word validness is a lot like syntactic correctness; the speaker may rely
+their knowledge of morphological rules to \textit{generate} new words. In fact
+Turkish has \textbf{recursive} morphology \citep{kabak2001} so a word can be
+inflected/derived in into infinitely many different ways. This is why
+computational processing of Turkish can never fully escape its morphology.
+
+\section{Conclusion}
+
+% Delete this todo
+\todo[inline]{Write a concluding summary. Also indicate what future work may be warranted. This section should be at least half a page in length.}
+
 \bibliographystyle{apalike}

### 2015-12-06 create/Main.hs
New sample and arrows in the HTML table

@@ -1,4 +1,5 @@
 {-# LANGUAGE UnicodeSyntax #-}
+{-# LANGUAGE BangPatterns  #-}
 
 module Main where
 
@@ -88,6 +89,13 @@ sample₆ = listArray (0, 25) [ "Cebren"
                             ,"."
                             ]
 
+sample₇ ∷ Array Int String
+sample₇ = listArray (0, 4) [ "Adam"
+                           , "sırıta"
+                           , "sırıta"
+                           , "yürüyordu"
+                           , "."]
+
 pretty ∷ Array Int String → [POS] → IO ()
 pretty ws ps = let lines = (\n → replicate n '-') <$> (length <$> ws)
                    align ∷ String → String → String
@@ -103,12 +111,15 @@ pretty ws ps = let lines = (\n → replicate n '-') <$> (length <$> ws)
                        zipWith align wsList pStrings
 
 table ∷ Array Int String → [POS] → String
-table ws ps = let makeItem w = "<th align=\"center\">" ++ w ++ "</th>"
+table ws ps = let len        = length ws
+                  makeItem w = "<th align=\"center\">" ++ w ++ "</th>"
                   makeRow xs = "<tr>" ++ concat (map makeItem xs) ++ "</tr>"
-                  wRow = makeRow $ foldr (:) [] ws
-                  pRow = makeRow $ map show ps
+                  wRow       = makeRow $ foldr (:) [] ws
+                  pRow       = makeRow $ map show ps
+                  arrows     = makeRow $ replicate len "↑"
               in    "<html><table>"
                  ++ wRow
+                 ++ arrows
                  ++ pRow
                  ++ "</table></html>"
 
@@ -153,7 +164,7 @@ main = do
                              n      = fromIntegral $ length taggedWordsList
                          in logFloat $ count' / n
       possibleTags     = [Noun .. Unknown]
-      newHMM           = HMM { states      = possibleTags ∷ [POS]
+      !newHMM           = HMM { states      = possibleTags ∷ [POS]
                              , events      = ws ∷ [String]
                              , initProbs   = initProbFn
                              , transMatrix = transFn
@@ -161,6 +172,7 @@ main = do
   writeFile "model.hmm" (show newHMM)
   -- putStrLn "Creating the model..."
   putStrLn $ table sample₆ $ viterbi newHMM sample₆
+  putStrLn $ table sample₇ $ viterbi newHMM sample₇
[…]

### 2015-12-07 create/Main.hs
Write and read model

@@ -4,7 +4,10 @@
 module Main where
 
 import           Data.Array           (Array, listArray)
-import           Data.HMM             (HMM (..), viterbi)
+import           Data.HMM             ( HMM (..)
+                                      , viterbi
+                                      , saveHMM
+                                      , loadHMM)
 import qualified Data.Map             as M
 import           Data.List            (intercalate)
 import           Data.Number.LogFloat
@@ -22,93 +25,6 @@ type 𝓛 = LogFloat -- Log-domain numbers to prevent underflow.
 ε = encodeFloat 1 $ fst  r - ds
   where r  = floatRange (0.1 :: Double)
         ds = floatDigits (0.1 :: Double)
-
-sample₁ ∷ Array Int String
-sample₁ = listArray (0, 4) [ "gözleri"
-                           , "kor"
-                           , "gibi"
-                           , "yanıyordu"
-                           , "."]
-
-sample₂ ∷ Array Int String
-sample₂ = listArray (0, 3) [ "adam"
-                           , "yine"
-                           , "geldi"
-                           , "."
-                           ]
-
-sample₃ ∷ Array Int String
-sample₃ = listArray (0, 4) [ "güzel"
-                           , "kız"
-                           , "mutlu"
-                           , "gözüküyordu"
-                           , "."]
-
-sample₄ ∷ Array Int String
-sample₄   = listArray (0, 5) [ "renksiz"
-                             , "yeşil"
-                             , "fikirler"
-                             , "sessizce"
-                             , "uyuyor"
-                             , "."
-                             ]
-
-sample₅ ∷ Array Int String
-sample₅   = listArray (0, 3) [ "dostlar"
-                             , "beni"
-                             , "hatırlasın"
-                             , "."
-                             ]
-
-sample₆ ∷ Array Int String
-sample₆ = listArray (0, 25) [ "Cebren"
-                            , "ve"
-                            , "hile"
-                            , "ile"
-                            , "aziz"
-                            , "vatanın"
-                            , ","
-                            , "bütün"
-                            , "kaleleri"
-                            , "zaptedilmiş"
-                            , "bütün"
-                            , "tersanelerine"
-                            , "girilmiş"
-                            , ","
-                            , "bütün"
-                            ,"orduları"
[…]

### 2015-12-07 tag/Main.hs

@@ -1,4 +1,7 @@
-{-# LANGUAGE BangPatterns #-}
+{-# LANGUAGE BangPatterns       #-}
+{-# LANGUAGE UnicodeSyntax      #-}
+{-# LANGUAGE StandaloneDeriving #-}
+{-# LANGUAGE FlexibleInstances  #-}
 
 module Main where
 
@@ -6,16 +9,140 @@ import Prelude
 import Data.HMM (loadHMM, viterbi, HMM(..))
 import System.IO
 import Data.Array
-import Parse hiding (words)
+import Parse (POS(..))
+import Data.Map.Strict (findWithDefault, Map(..))
+import Data.List (intercalate)
+import Data.Number.LogFloat
+
+type ℤ = Int
+type ℚ = Double
+type 𝓛 = LogFloat -- Log-domain numbers to prevent underflow.
+
+probability ∷ (Ord a, Ord b) ⇒ (a, b) → Map a ℤ → Map (a, b) ℤ → 𝓛
+probability (x, y) c₁ c₂ = if xCount == 0 || yCount == 0
+                           then logFloat ε
+                           else (logFloat yCount) / (logFloat xCount)
+  where yCount = fromIntegral (findWithDefault 0 (x, y) c₂) ∷ ℚ
+        xCount = fromIntegral (findWithDefault 0 x c₁) ∷ ℚ
+
+ε ∷ Double
+ε = encodeFloat 1 $ fst  r - ds
+  where r  = floatRange (0.1 :: Double)
+        ds = floatDigits (0.1 :: Double)
+
+
+sample₁ ∷ Array Int String
+sample₁ = listArray (0, 4) [ "gözleri"
+                           , "kor"
+                           , "gibi"
+                           , "yanıyordu"
+                           , "."]
+
+sample₂ ∷ Array Int String
+sample₂ = listArray (0, 3) [ "adam"
+                           , "yine"
+                           , "geldi"
+                           , "."
+                           ]
+
+sample₃ ∷ Array Int String
+sample₃ = listArray (0, 4) [ "güzel"
+                           , "kız"
+                           , "mutlu"
+                           , "gözüküyordu"
+                           , "."]
+
+sample₄ ∷ Array Int String
+sample₄   = listArray (0, 5) [ "renksiz"
+                             , "yeşil"
+                             , "fikirler"
+                             , "sessizce"
+                             , "uyuyor"
+                             , "."
+                             ]
+
+sample₅ ∷ Array Int String
[…]

### 2015-12-07 create/Main.hs
Refactor to avoid warnings

@@ -3,13 +3,8 @@
 
 module Main where
 
-import           Data.Array           (Array, listArray)
-import           Data.HMM             ( HMM (..)
-                                      , viterbi
-                                      , saveHMM
-                                      , loadHMM)
+import           Data.Array           (Array)
 import qualified Data.Map             as M
-import           Data.List            (intercalate)
 import           Data.Number.LogFloat
 import           Parse
 import           Prelude              hiding (putStr, words, lines)
@@ -73,19 +68,8 @@ main = do
       tagBigrams       = [(ts !! i, ts !! (i+1)) | i ← [0 .. (length ts)-2]]
       tagBigramFreqs   = freqMap tagBigrams
       -- We will use these to create our HMM.
-      transFn s₁ s₂     = probability (s₁, s₂) tagFreqs tagBigramFreqs
-      outFn s e        = probability (e, s) wordFreqs taggedWordFreqs
       initStatesFreqs  = freqMap  $ map (head . map snd) taggedWordsList
-      initProbFn s     = let count  = M.findWithDefault 0 s initStatesFreqs
-                             count' = fromIntegral count
-                             n      = fromIntegral $ length taggedWordsList
-                         in logFloat $ count' / n
-      possibleTags     = [Noun .. Unknown]
-      !newHMM           = HMM { states      = possibleTags ∷ [POS]
-                             , events      = ws ∷ [String]
-                             , initProbs   = initProbFn
-                             , transMatrix = transFn
-                             , outMatrix   = outFn}
+  createDirectoryIfMissing False "model"
   writeFile "model/tagFreqs.hs"        (show tagFreqs)
   writeFile "model/wordFreqs.hs"       (show wordFreqs)
   writeFile "model/tagBigramFreqs.hs"  (show tagBigramFreqs)

### 2015-12-07 tag/Main.hs

@@ -5,12 +5,11 @@
 
 module Main where
 
-import Prelude
-import Data.HMM (loadHMM, viterbi, HMM(..))
-import System.IO
+import Prelude hiding (lines)
+import Data.HMM (viterbi, HMM(..))
 import Data.Array
 import Parse (POS(..))
-import Data.Map.Strict (findWithDefault, Map(..))
+import Data.Map.Strict (findWithDefault, Map)
 import Data.List (intercalate)
 import Data.Number.LogFloat
 

### 2015-12-07 create/Main.hs
Clean things up

@@ -3,13 +3,11 @@
 
 module Main where
 
-import           Data.Array           (Array)
 import qualified Data.Map             as M
 import           Data.Number.LogFloat
 import           Parse
 import           Prelude              hiding (putStr, words, lines)
 import           System.Directory
--- import           System.IO (putStr)
 import           Text.XML.HXT.Core
 
 type ℤ = Int      -- The integer type `Int` will be denoted by ℤ.
@@ -20,19 +18,6 @@ type 𝓛 = LogFloat -- Log-domain numbers to prevent underflow.
 ε = encodeFloat 1 $ fst  r - ds
   where r  = floatRange (0.1 :: Double)
         ds = floatDigits (0.1 :: Double)
-
-table ∷ Array Int String → [POS] → String
-table ws ps = let len        = length ws
-                  makeItem w = "<th align=\"center\">" ++ w ++ "</th>"
-                  makeRow xs = "<tr>" ++ concat (map makeItem xs) ++ "</tr>"
-                  wRow       = makeRow $ foldr (:) [] ws
-                  pRow       = makeRow $ map show ps
-                  arrows     = makeRow $ replicate len "↑"
-              in    "<html><table>"
-                 ++ wRow
-                 ++ arrows
-                 ++ pRow
-                 ++ "</table></html>"
 
 freqMap ∷ Ord a ⇒ [a] → M.Map a ℤ
 freqMap unigrams = populate M.empty unigrams
@@ -76,15 +61,4 @@ main = do
   writeFile "model/taggedWordFreqs.hs" (show taggedWordFreqs)
   writeFile "model/initStatesFreqs.hs" (show initStatesFreqs)
   writeFile "model/words.hs" (show ws)
-  return ()
-  -- putStrLn "Creating the model..."
-  -- putStr "\n"
-  -- pretty sample₂ $ viterbi newHMM sample₂
-  -- putStr "\n"
-  -- pretty sample₃ $ viterbi newHMM sample₃
-  -- putStr "\n"
-  -- pretty sample₄ $ viterbi newHMM sample₄
-  -- putStr "\n"
-  -- pretty sample₅ $ viterbi newHMM sample₅
-  -- putStr "\n"
-  -- pretty sample₆ $ viterbi newHMM sample₆
+  putStrLn "Successfully saved model in model."

### 2015-12-07 tag/Main.hs

@@ -3,7 +3,7 @@
 {-# LANGUAGE StandaloneDeriving #-}
 {-# LANGUAGE FlexibleInstances  #-}
 
-module Main where
+module Main (main) where
 
 import Prelude hiding (lines)
 import Data.HMM (viterbi, HMM(..))
@@ -117,6 +117,19 @@ pretty ws ps = let lines = (\n → replicate n '-') <$> (length <$> ws)
                      putStrLn $ intercalate "  " $
                        zipWith align wsList pStrings
 
+table ∷ Array Int String → [POS] → String
+table ws ps = let len        = length ws
+                  makeItem w = "<th align=\"center\">" ++ w ++ "</th>"
+                  makeRow xs = "<tr>" ++ concat (map makeItem xs) ++ "</tr>"
+                  wRow       = makeRow $ foldr (:) [] ws
+                  pRow       = makeRow $ map show ps
+                  arrows     = makeRow $ replicate len "↑"
+              in    "<html><table>"
+                 ++ wRow
+                 ++ arrows
+                 ++ pRow
+                 ++ "</table></html>"
+
 main :: IO ()
 main = do
   tagFreqsStr        ← readFile "model/tagFreqs.hs"
@@ -144,4 +157,15 @@ main = do
                             , transMatrix = transFn
                             , outMatrix   = outFn
                             }
-    in pretty sample₁ $ viterbi newHMM sample₁
+    in do
+      putStrLn $ table sample₆ (viterbi newHMM sample₆)
+      putStrLn $ "<hr>"
+      putStrLn $ table sample₅ (viterbi newHMM sample₅)
+      putStrLn $ "<hr>"
+      putStrLn $ table sample₄ (viterbi newHMM sample₄)
+      putStrLn $ "<hr>"
+      putStrLn $ table sample₃ (viterbi newHMM sample₃)
+      putStrLn $ "<hr>"
+      putStrLn $ table sample₂ (viterbi newHMM sample₂)
+      putStrLn $ "<hr>"
+      putStrLn $ table sample₁ (viterbi newHMM sample₁)

### 2015-12-07 paper/paper.tex
New sample and arrows in the HTML table
Write about performance

--- 2015-11-17 proposal/proposal.tex
@@
 morphology, its syntax seems to be a less interesting problem.
-\section{Proposal}
+\section{Proposed Work}
 
@@
   most of the cases. There are certain exceptions to this, such as
-  reduplication which I denote with the POS tag ``Dup''} and will not be accounting for morphology, unlike
+  reduplication which I denote with the POS tag ``Dup''} and will not be
+accounting for morphology, unlike
 the case with most of the serious research on Turkish NLP. There definitely are
@@
 
+\section{Results}
+The POS tagger I've built as proposed can successfully tag words with a
+reasonable accuracy.
+
+\subsection{Performance}
+Parsing the entire corpus and creating a model out of it takes 10.5 seconds on
+average. Tagging a sentence (which we're assuming to have 5 words on average)
+takes $~0.3$ seconds. The implementation of the \texttt{Data.HMM} datatype uses
+\texttt{Data.Array}'s
+\footnote{https://hackage.haskell.org/package/array-0.5.1.0/docs/Data-Array.html}
+which is why it runs reasonably fast.
+
+\subsection{Limitations}
+There is a fundamental limitation to the POS tagger I built: Turkish morphology
+is essential to the overall completeness of Turkish.
+
+In English we entertain the notion of \emph{syntactic correctness}, which is
+more or less a formalism of correct ordering of words. A speaker of English that has
+synactic competence in the language \citep{bachman1990} may know that a sentence
+is well-formed, relying on his knowledge of syntactic rules. However, for the
+\textit{words}, knowledge of morphological rules is not adequate to conclude
+that a word is made up of correctly ordered morphemes; they have to rely on
+their lexical knowledge to verify that the morphological form they
+derived/inflected is a valid one. Consider the word ``satisfiableness'' which
+is not a valid word according to several dictionaries, although being derived
+with valid suffixes. The correct form of this concept is ``satisfiability''.
+
+In Turkish, the concept of word validness is a lot like syntactic correctness;
+the speaker may rely their knowledge of morphological rules to \emph{generate}
+new words. In fact Turkish has \textbf{recursive} morphology \citep{kabak2001}
+so a word can be inflected/derived in into infinitely many different forms. This
+is why computational processing of Turkish can never fully escape its
+morphology.
+
+The hidden markov model I've built for this project \emph{ignores} morphology.
+
+\section{Conclusion}
+
+% Delete this todo
+\todo[inline]{Write a concluding summary. Also indicate what future work may be warranted. This section should be at least half a page in length.}
+
 \bibliographystyle{apalike}

### 2015-12-08 paper/paper.tex
Comment on some of the code

@@
 \usepackage{hyperref}
+\usepackage{listings}
+\usepackage{amsmath}
+\usepackage[T1]{fontenc}
+\usepackage{mathrsfs}
 
@@
 \begin{document}
+%include agda.fmt
 \maketitle
@@
 
+\subsection{A Brief Overview on Implementation}
+The project consists of two executables: (1) \texttt{create-model} that reads
+the data from the corpus and parses it into a form from which we can easily
+construct an \texttt{HMM}. Unfortunately, it is highly impractical---if not
+impossible---to directly serialize \texttt{HMM} since it has three fields
+containing functions
+\footnote{http://stackoverflow.com/questions/17785916/can-haskell-functions
+  -be-serialized for further information.}. We consider this a fault in the
+design of the \texttt{Data.HMM} package and urge further development. Although
+this has almost no effect on performance, it obfuscates the code.
+
+Since processing the corpus is a resource-demanding operation,
+\texttt{create-model} is intended be run only once after which \texttt{run-tag}
+can be used to make use of that data.
+
+\paragraph{\texttt{create-model}}
+The executable \texttt{create-model} is implemented in two files:
+\texttt{Parse.hs} and \texttt{Main.hs}. \texttt{Parse.hs} handles the parsing of
+the corpus.
+
+The function \texttt{freqMap} takes as input a list of orderable elements
+(denoted with \texttt{Ord a}) and creates a \texttt{Map} containing, as value,
+the frequency of each element.
+
+\begin{lstlisting}[
+  language=Haskell,
+  mathescape,
+  columns=fullflexible,
+  basicstyle=\small\ttfamily
+]
+freqMap :: Ord a $\Rightarrow$ [a] $\rightarrow$ M.Map a Int
+freqMap unigrams = populate M.empty unigrams
+  where populate :: Ord a M.Map a $Z$ $\rightarrow$ [a] $\rightarrow$ M.Map a Int
+        populate m [] = m
+        populate m (x:xs) = let freq = (M.findWithDefault 0 x m) :: ℤ
+                            in populate (M.insert x (freq + 1) m) xs
+\end{lstlisting}
+
+This is the basis of our \emph{supervised} learning.
+
+\paragraph{\texttt{tag}}
+
+The executable \texttt{run-tag} is where the actual tagging happens. The
+frequencies we have pre-computed in \texttt{create-model} are passed in to the
+probability function which we give below.
+\begin{lstlisting}[
+  language=Haskell,
+  mathescape,
+  columns=fullflexible,
+  basicstyle=\small\ttfamily
+]
+probability :: (Ord a, Ord b) $\Rightarrow$ (a, b) $\rightarrow$ Map a Int$\rightarrow$ Map (a, b) ℤ $\rightarrow \mathscr{L}$
+probability (x, y) c$\_1$ c$\_2$ = if xCount == 0 || yCount == 0
+                           then logFloat $\epsilon$
+                           else (logFloat yCount) / (logFloat xCount)
+  where yCount = fromIntegral (findWithDefault 0 (x, y) c₂) ∷ Double
+        xCount = fromIntegral (findWithDefault 0 x c₁) ∷ Double
+\end{lstlisting}
 \subsection{Performance}
@@
 takes $~0.3$ seconds. The implementation of the \texttt{Data.HMM} datatype uses
-\texttt{Data.Array}'s
+\texttt{Data.Array}
 \footnote{https://hackage.haskell.org/package/array-0.5.1.0/docs/Data-Array.html}

### 2015-12-08 paper/Main.hs

@@ -1,0 +1,58 @@
+{-# LANGUAGE UnicodeSyntax #-}
+{-# LANGUAGE BangPatterns  #-}
+
+module Main where
+
+import qualified Data.Map             as M
+import           Data.Number.LogFloat
+import           Parse
+import           Prelude              hiding (putStr, words, lines)
+import           System.Directory
+import           Text.XML.HXT.Core
+
+type ℤ = Int      -- The integer type `Int` will be denoted by ℤ.
+type ℚ = Double   -- The double type will be denoted by ℚ.
+type 𝓛 = LogFloat -- Log-domain numbers to prevent underflow.
+
+ε ∷ Double
+ε = encodeFloat 1 $ fst  r - ds
+  where r  = floatRange (0.1 :: Double)
+        ds = floatDigits (0.1 :: Double)
+
+freqMap ∷ Ord a ⇒ [a] → M.Map a ℤ
+freqMap unigrams = populate M.empty unigrams
+  where populate ∷ Ord a ⇒ M.Map a ℤ → [a] → M.Map a ℤ
+        populate m [] = m
+        populate m (x:xs) = let freq = (M.findWithDefault 0 x m) ∷ ℤ
+                            in populate (M.insert x (freq + 1) m) xs
+
+--  | Takes in a tuple (x, y) (i.e., a bigram), and takes in two frequency maps
+--    c₁ and c₂, c₁ for counting all the occurrences of x and the other for
+--    counting the number of bigrams.
+
+main ∷ IO ()
+main = do
+  files ← getDirectoryContents "tb_uni"
+  let fileNames = drop 2 . take 1500 $ map ("tb_uni/" ++) files
+  pairList ← mapM runX $ map getWords fileNames
+  let taggedWordsList  = map parseTupleList pairList
+      taggedWords ∷ [(String, POS)]
+      taggedWords      = concat taggedWordsList
+      (ws, ts)         = unzip taggedWords
+      -- For computing the probability of word-tag pair.
+      taggedWordFreqs  = freqMap taggedWords
+      wordFreqs        = freqMap ws
+      tagFreqs         = freqMap ts
+      -- For computing the probability of tag bigram
+      tagBigrams       = [(ts !! i, ts !! (i+1)) | i ← [0 .. (length ts)-2]]
+      tagBigramFreqs   = freqMap tagBigrams
+      -- We will use these to create our HMM.
[…]

### 2015-12-08 paper/Main.hs
Delete

@@ -1,58 +1,0 @@
-{-# LANGUAGE UnicodeSyntax #-}
-{-# LANGUAGE BangPatterns  #-}
-
-module Main where
-
-import qualified Data.Map             as M
-import           Data.Number.LogFloat
-import           Parse
-import           Prelude              hiding (putStr, words, lines)
-import           System.Directory
-import           Text.XML.HXT.Core
-
-type ℤ = Int      -- The integer type `Int` will be denoted by ℤ.
-type ℚ = Double   -- The double type will be denoted by ℚ.
-type 𝓛 = LogFloat -- Log-domain numbers to prevent underflow.
-
-ε ∷ Double
-ε = encodeFloat 1 $ fst  r - ds
-  where r  = floatRange (0.1 :: Double)
-        ds = floatDigits (0.1 :: Double)
-
-freqMap ∷ Ord a ⇒ [a] → M.Map a ℤ
-freqMap unigrams = populate M.empty unigrams
-  where populate ∷ Ord a ⇒ M.Map a ℤ → [a] → M.Map a ℤ
-        populate m [] = m
-        populate m (x:xs) = let freq = (M.findWithDefault 0 x m) ∷ ℤ
-                            in populate (M.insert x (freq + 1) m) xs
-
---  | Takes in a tuple (x, y) (i.e., a bigram), and takes in two frequency maps
---    c₁ and c₂, c₁ for counting all the occurrences of x and the other for
---    counting the number of bigrams.
-
-main ∷ IO ()
-main = do
-  files ← getDirectoryContents "tb_uni"
-  let fileNames = drop 2 . take 1500 $ map ("tb_uni/" ++) files
-  pairList ← mapM runX $ map getWords fileNames
-  let taggedWordsList  = map parseTupleList pairList
-      taggedWords ∷ [(String, POS)]
-      taggedWords      = concat taggedWordsList
-      (ws, ts)         = unzip taggedWords
-      -- For computing the probability of word-tag pair.
-      taggedWordFreqs  = freqMap taggedWords
-      wordFreqs        = freqMap ws
-      tagFreqs         = freqMap ts
-      -- For computing the probability of tag bigram
-      tagBigrams       = [(ts !! i, ts !! (i+1)) | i ← [0 .. (length ts)-2]]
-      tagBigramFreqs   = freqMap tagBigrams
-      -- We will use these to create our HMM.
[…]

### 2015-12-08 create/Main.hs
Move the probability function to `tag`

@@ -29,12 +29,6 @@ freqMap unigrams = populate M.empty unigrams
 --  | Takes in a tuple (x, y) (i.e., a bigram), and takes in two frequency maps
 --    c₁ and c₂, c₁ for counting all the occurrences of x and the other for
 --    counting the number of bigrams.
-probability ∷ (Ord a, Ord b) ⇒ (a, b) → M.Map a ℤ → M.Map (a, b) ℤ → 𝓛
-probability (x, y) c₁ c₂ = if xCount == 0 || yCount == 0
-                           then logFloat ε
-                           else (logFloat yCount) / (logFloat xCount)
-  where yCount = fromIntegral (M.findWithDefault 0 (x, y) c₂) ∷ ℚ
-        xCount = fromIntegral (M.findWithDefault 0 x c₁) ∷ ℚ
 
 main ∷ IO ()
 main = do

### 2015-12-08 create/Parse.hs

@@ -67,7 +67,7 @@ parseTupleList xs = map parseTuple xs
                 info' = extractPOS info
 
 -- | Given number of files `numFiles` takes out all word-tag pairs
--- | from each file indexed from 0 .. `numFiles`, and concatenates
--- | them.
+--   from each file indexed from 0 .. `numFiles`, and concatenates
+--   them.
 getWords :: FilePath -> IOSLA (XIOState s) a (String, String)
 getWords f = readDocument [withValidate no] f >>> words

### 2015-12-08 paper/paper.tex
Write

@@
 \title{\bf An HMM-based part-of-speech tagger for Turkish}
-\author{Ayberk Tosun}
+\author{Ayberk Tosun\\\texttt{tosun2@illinois.edu}}
 \date{}
@@
 further development. Moreover, \citet{Korkut2015} began the development of a
-Turkish NLP framework named ``Guguk,'' implemented in the programming language Haskell. I intend
+Turkish NLP library named ``Guguk'', implemented in the programming language Haskell. I intend
 to merge this project to Guguk, hence contributing towards
-the existence of a Turkish NLP-framework in Haskell.
+the existence of a Turkish NLP-library in Haskell.
+
+Before proceeding, a few remarks on notational conventions: I use
+\texttt{HMM} (with the fixed-width font) to the datatype I'm using from the
+\texttt{Data.HMM} package and HMM (with regular font) to refer to the concept.
 
@@
 % least one to two paragraphs in length.
-The most important advancement in the computational linguistics of Turkish is
-probably the NLP framework ``Zemberek'' by \citet{akin2007zemberek}, which has been
-immensely useful for both the theory and practice of Turkish NLP. Although it started
-out as an NLP tool for the Turkish language, Zemberek moved on to become a
-a tool for processing all Turkic languages.
+\cite{dincer2008suffix} built a POS tagger for Turkish using HMMs on suffixes.
+This proves an effective approach seeing that (1) Turkish morphology is
+exclusively based on suffixes (not prefixes) (2) morphology is central to the
+Turkish language so it is likely that any given word is formed through the
+agglutination of at least one morpheme. The tagger described has ``\%90.2
+accuracy in the best case''.
 
 The widely cited \citet{oflazer1994tagging} outlines the implementation of a
-tool for both morpheme glossing and POS tagging, that has ``98-99 \%'' accuracy. This
-tool is based on a morphological specification of Turkish and makes use of a
-morphological processor called ``PC-Kimmo'' \citep{antworth1991pc}.
+tool for both morpheme glossing and POS tagging, that has ``98-99 \%'' accuracy.
+This tool is based on a morphological specification of Turkish and makes use of
+a morphological parser called ``PC-Kimmo'' \citep{antworth1991pc}. I should
+stress that \citet{oflazer1994tagging} does POS tagging as a by-product of its
+capability of morphological parsing.
 
-In general, most of the literature relating to the computational processing of Turkish
-pertain primarily to morphology---compared to the complexity of Turkish
+In general, most of the literature relating to the computational processing of
+Turkish pertain primarily to morphology. Compared to the complexity of Turkish
 morphology, its syntax seems to be a less interesting problem.
+
 \section{Proposed Work}
 
-I will implement a part-of-speech tagger for the Turkish language. As
-mentioned in section~\ref{sec:related_work}, I will mostly focus on parts-of-speech
-of words\footnote{I base my definition of a ``word'' on the orthographic space in
-  most of the cases. There are certain exceptions to this, such as
-  reduplication which I denote with the POS tag ``Dup''} and will not be
-accounting for morphology, unlike
-the case with most of the serious research on Turkish NLP. There definitely are
-problems with this approach---but fixing these would go beyond the scope of
-this class
+I will implement a part-of-speech tagger for the Turkish language. As mentioned
+in section~\ref{sec:related_work}, I will mostly focus on parts-of-speech of
+words\footnote{I base my definition of a ``word'' on the orthographic space in
+  most of the cases. There are certain exceptions to this, such as duplication
+  which I denote with the POS tag ``Dup''} and will not be accounting for
+morphology, unlike most of the related work I've mentioned. There are definite
+problems with this approach. Fixing these would go beyond the scope of this
+paper therefore I defer the integration of morphology for further development.
 
-For programming the POS tagger I will be using Haskell, a statically-typed and purely
-functional programming language, along with the
-\href{\hmmURL}{\texttt{Data.HMM}} package.
-The data, that is probably the most important component of this
-project, comes from the METU-Sabanc{\i} treebank built by
-\citet{oflazer2003building}.  I am planning to strip the POS tag for each word from
-the treebank and use those to train the HMM
+For programming the POS tagger I will be using Haskell, a statically-typed and
+purely functional programming language, along with the
+\href{\hmmURL}{\texttt{Data.HMM}} package. The data, that is probably the most
+important component of this project, comes from the METU-Sabanc{\i} treebank
+built by \citet{oflazer2003building}. I am planning to strip the POS tag for
+each word from the treebank and use those to train the HMM
 
 \section{Results}
-The POS tagger I've built as proposed can successfully tag words with a
-reasonable accuracy.
+The POS tagger I've built as proposed can tag words with a
+reasonable accuracy which we talk about in \ref{subsec:accuracy}.
 
-\subsection{A Brief Overview on Implementation}
+\subsection{A Brief Overview of the Program}
 The project consists of two executables: (1) \texttt{create-model} that reads
@@
 \end{lstlisting}
-
-This is the basis of our \emph{supervised} learning.
+Once the required lists are generated using these they are saved to a file in
+the directory \texttt{model}.
 
@@
 ]
-probability :: (Ord a, Ord b) $\Rightarrow$ (a, b) $\rightarrow$ Map a Int$\rightarrow$ Map (a, b) ℤ $\rightarrow \mathscr{L}$
-probability (x, y) c$\_1$ c$\_2$ = if xCount == 0 || yCount == 0
+probability :: (Ord a, Ord b) $\Rightarrow$ (a, b) $\rightarrow$ Map a Int $\rightarrow$ Map (a, b) ℤ $\rightarrow \mathscr{L}$
+probability (x, y) c$_1$ c$_2$ = if xCount == 0 || yCount == 0
                                    then logFloat $\epsilon$
@@
 \end{lstlisting}
+
+\texttt{probability} when combined with \texttt{freqMap} yield an efficient
+basis for implementing Bayes' law. When we are creating the functions
+required by \texttt{HMM} we are storing the map returned by a call to
+\texttt{freqMap} in the lexical closure of \texttt{probability} by making use
+Haskell's currying so that we can eloquently define the two crucial functions
+as follows:
+\begin{lstlisting}[
+  language=Haskell,
+  mathescape,
+  columns=fullflexible,
+  basicstyle=\small\ttfamily
+]
+transFn s$_1$ s$_2$ = probability (s$_1$, s$_2$) tagFreqs tagBigramFreqs
+outFn s e = probability (e, s) wordFreqs taggedWordFreqs
+\end{lstlisting}
+
+\subsection{Accuracy}
+\label{subsec:accuracy}
+
+I have tested the POS tagger with 6075 words from a small portion of the corpus
+that I reserved for testing (i.e., did not use for tranining the HMM). The POS
+tagger tagged 4514 of these words correctly (74.3 \%). Here is a brief list of
+known problems:
+\begin{description}
+  \item[Duplications:] The Turkish language contains lots of duplications.
+  Syntactically speaking, a duplication behaves like an word itself. In the
+  treebank, reduplication are denoted as as single word composed of joining the
+  two halves of the duplication with an underscore (half$_1$\_$half_2$). Since
+  there is not an easy way to detect whether two sequential words are
+  reduplications or not, I am giving duplications in an incorrect format to the
+  HMM. I am predicting (based on my informal tests with the model) that
+  correction of this problem would increase the accuracy by about $\%5$.
+  \item[Relaxed word order:] Turkish is generally considered a free constituent
+  order language \citep{erguvanli, dincer2008suffix}, although there does not seem to be an
+  agreed-upon view on this. Yet, it would be safe to say that the word order of
+  Turkish is more relaxed compared to English. The HMM-based approach to POS
+  tagging presupposes the ordered-ness of the language. Although there is a
+  canonical word order in Turkish, it tends not to be rigidly followed all the
+  time---the morphosyntax of Turkish allows variation. It might be possible to
+  increase the accuracy of the HMM by, for example, isolating bigrams in
+  dependent clauses from those in their parents might work since Turkish, due to
+  its relaxed order, permits lots of dependent clauses.
+\end{description}
+
 \subsection{Performance}
 Parsing the entire corpus and creating a model out of it takes 10.5 seconds on
-average. Tagging a sentence (which we're assuming to have 5 words on average)
+average. Tagging a sentence (which we are assuming to have 5 words on average)
 takes $~0.3$ seconds. The implementation of the \texttt{Data.HMM} datatype uses

### 2015-12-08 create/Main.hs
Chang message

@@ -55,4 +55,4 @@ main = do
   writeFile "model/taggedWordFreqs.hs" (show taggedWordFreqs)
   writeFile "model/initStatesFreqs.hs" (show initStatesFreqs)
   writeFile "model/words.hs" (show ws)
-  putStrLn "Successfully saved model in model."
+  putStrLn "Successfully saved model in dir \"model\"."

### 2015-12-09 paper/paper.tex
Almost complete

@@
 \section{Results}
-The POS tagger I've built as proposed can tag words with a
-reasonable accuracy which we talk about in \ref{subsec:accuracy}.
+The resulting program I have implemented is based on bigram hidden markov
+models. To summarize it we are making use of three probabilities: (1) $P(t_i |
+t_{i-1})$ the probability of being at a state (i.e., a tag) given the previous
+state. This is computed\footnote{Although I use the word ``compute'', everything
+  is already pre-computed and the referred computation is merely a lookup. This
+  is discussed in \ref{par:tag}} by the function \texttt{transFn}. (2) $P(w_i | t_i)$ the
+probability of seeing a word having a given tag, computed by the function
+\texttt{outFn}. Finally, the probability of a given state being the start state
+of a sentence, computed by the function \texttt{outProbs}.
 
@@
 
-\paragraph{\texttt{create-model}}
+\subsubsection{\texttt{create-model}}
 The executable \texttt{create-model} is implemented in two files:
@@
 freqMap :: Ord a $\Rightarrow$ [a] $\rightarrow$ M.Map a Int
-freqMap unigrams = populate M.empty unigrams
-  where populate :: Ord a M.Map a $Z$ $\rightarrow$ [a] $\rightarrow$ M.Map a Int
+freqMap elems = populate empty elems
+  where populate :: Ord a Map a Int $\rightarrow$ [a] $\rightarrow$ M.Map a Int
          populate m [] = m
-        populate m (x:xs) = let freq = (M.findWithDefault 0 x m) :: ℤ
+         populate m (x:xs) = let freq = (M.findWithDefault 0 x m) :: Int
                             in populate (M.insert x (freq + 1) m) xs
 \end{lstlisting}
-Once the required lists are generated using these they are saved to a file in
-the directory \texttt{model}.
 
-\paragraph{\texttt{tag}}
+This function is called several times with lists of values we want to
+\emph{count} such as tag bigrams, a given word with a certain tag et cetera.
+Pre-computing the counts is almost mandatory since frequencies (i.e., counts)
+will need to be accessed frequently when computing the probabilities. Once the
+necessary maps are generated, they are written to files in the directory
+\texttt{model}.
 
+\subsubsection{\texttt{tag}}
+\label{par:tag}
 The executable \texttt{run-tag} is where the actual tagging happens. The
@@
 
-\texttt{probability} when combined with \texttt{freqMap} yield an efficient
-basis for implementing Bayes' law. When we are creating the functions
+This is a straightforward implementation of probability relying on a given count
+map. One thing to note is that it returns $\epsilon$ probability (i.e., the
+smalles value a \texttt{logFloat}) if a given element is not found.
+
+\texttt{probability} when combined with \texttt{freqMap} forms an
+efficient method for implementing Bayes' law. When we are creating the functions
 required by \texttt{HMM} we are storing the map returned by a call to
 \texttt{freqMap} in the lexical closure of \texttt{probability} by making use
-Haskell's currying so that we can eloquently define the two crucial functions
-as follows:
+Haskell's currying so that we can eloquently define the two crucial functions as
+follows:
 \begin{lstlisting}[
@@
 \end{lstlisting}
+The probability functions used to create \texttt{transFn} and \texttt{outFn} hold
+the count maps \texttt{tagBigramFreqs} and \texttt{taggedWordFreqs} in their
+lexical closure, respectively.
 
-\subsection{Accuracy}
-\label{subsec:accuracy}
+Finally we declare a new \texttt{HMM} as follows:
 
-I have tested the POS tagger with 6075 words from a small portion of the corpus
-that I reserved for testing (i.e., did not use for tranining the HMM). The POS
-tagger tagged 4514 of these words correctly (74.3 \%). Here is a brief list of
-known problems:
-\begin{description}
-  \item[Duplications:] The Turkish language contains lots of duplications.
-  Syntactically speaking, a duplication behaves like an word itself. In the
-  treebank, reduplication are denoted as as single word composed of joining the
-  two halves of the duplication with an underscore (half$_1$\_$half_2$). Since
-  there is not an easy way to detect whether two sequential words are
-  reduplications or not, I am giving duplications in an incorrect format to the
-  HMM. I am predicting (based on my informal tests with the model) that
-  correction of this problem would increase the accuracy by about $\%5$.
-  \item[Relaxed word order:] Turkish is generally considered a free constituent
-  order language \citep{erguvanli, dincer2008suffix}, although there does not seem to be an
-  agreed-upon view on this. Yet, it would be safe to say that the word order of
-  Turkish is more relaxed compared to English. The HMM-based approach to POS
-  tagging presupposes the ordered-ness of the language. Although there is a
-  canonical word order in Turkish, it tends not to be rigidly followed all the
-  time---the morphosyntax of Turkish allows variation. It might be possible to
-  increase the accuracy of the HMM by, for example, isolating bigrams in
-  dependent clauses from those in their parents might work since Turkish, due to
-  its relaxed order, permits lots of dependent clauses.
-\end{description}
+\newpage
+\begin{lstlisting}[
+  language=Haskell,
+  mathescape,
+  columns=fullflexible,
+  basicstyle=\small\ttfamily
+]
+      newHMM = HMM { states      = possibleTags ∷ [POS]
+                     , events      = ws ∷ [String]
+                     , initProbs   = initProbFn
+                     , transMatrix = transFn
+                     , outMatrix   = outFn
+                     }
+\end{lstlisting}
+
+This is a simple record (a record is like C struct in Haskell) containing the
+probability functions we defined.\footnote{For some reason, the author of the
+  \texttt{Data.HMM} library named the fields (i.e., attribute), that are of
+  function types, using names that end with \texttt{-Matrix} although they are
+  \emph{not} matrices. The reader should not be confused by this.}
 
@@
 
+\section{Conclusion}
+
+We have shown that it is possible---though not ideal---to build a POS tagger for
+Turkish that is based entirely on a bigram hidden markov model. Let us first
+look at the accuracy of this model and propose possible fixes that do not
+fundamentally change the structure of the model.
+
+\subsection{Accuracy}
+\label{subsec:accuracy}
+
+We have tested the POS tagger with 6075 words from a small portion of the corpus
+that I reserved for testing (i.e., did not use for tranining the HMM). The POS
+tagger tagged 4514 of these words correctly (74.3 \%). Here is a brief list of
+known problems:
+
+\begin{description}
+
+  \item[Duplications:] The Turkish language contains many duplications.
+  Syntactically speaking, a duplication behaves like a word itself. In the
+  treebank, duplications are denoted with two halves joint the with an
+  underscore. Since there is not an easy way to detect whether two sequential
+  words are reduplications or not, we are giving duplications in an incorrect
+  format to the HMM.  We predict that (based on informal tests with the model)
+  that correction of this problem would significantly increase the accuracy.
+
+  \item[Abundance of independent clauses:] It is common in the Turkish language
+  to have multiple independent clauses in a given sentence. It is possible that
+  this decreases the accuracy of the HMM: instead of a tag $t$ being followed by
+  a tag $t'$, we get a lot of $t, <clause>, t'$ which obscures the word order we
+  are interested in. Since we may always assign the \textsc{Punc} tag to a comma
+  with full certainty, we may simply ignore commas as far as probability is
+  concerned and tag them with \textsc{Punc}.
+
+\end{description}
+
 \subsection{Limitations}
@@
 In English we entertain the notion of \emph{syntactic correctness}, which is
-more or less a formalism of correct ordering of words. A speaker of English that has
-synactic competence in the language \citep{bachman1990} may know that a sentence
-is well-formed, relying on his knowledge of syntactic rules. However, for the
-\textit{words}, knowledge of morphological rules is not adequate to conclude
-that a word is made up of correctly ordered morphemes; they have to rely on
-their lexical knowledge to verify that the morphological form they
-derived/inflected is a valid one. Consider the word ``satisfiableness'' which
-is not a valid word according to several dictionaries, although being derived
-with valid suffixes. The correct form of this concept is ``satisfiability''.
+more or less a formalism of correct ordering of words. A speaker of English that
+has synactic competence in the language \citep{bachman1990} may know that a
+sentence is well-formed, relying on his knowledge of syntactic rules. However,
+for the \textit{words}, knowledge of morphological rules is not adequate to
+conclude that a word is made up of correctly ordered morphemes; they have to
+rely on their lexical knowledge to verify that the morphological form they
+derived/inflected is a valid one. Consider the word ``satisfiableness'' which is
+not a valid word according to several dictionaries, although being derived with
+valid suffixes. The correct form of this concept is ``satisfiability''.
 
@@
 
-The hidden markov model I've built for this project \emph{ignores} morphology.
+\subsection{Further research}
 
-\section{Conclusion}
-
-% Delete this todo
-\todo[inline]{Write a concluding summary. Also indicate what future work may be warranted. This section should be at least half a page in length.}
+A salient possibility for further research is the incorporation of word endings
+into the model using regular expressions. It is trivial to express Turkish
+suffixes, that change their form as not to upset the major vowel harmony in a
+given word. We may say with probably more than $90\%$ probability that a given
+word has a certain POS looking whether its ending satisfies regular expressions.
 

### 2015-12-09 paper/paper.tex
Write some more

@@
 \label{sec:motivation}
-My main motivation for undertaking this project is to understand the
-principles underlying POS-tagging. Due to time and resource constraints, it is
-not possible for me to build a state-of-the-art POS-tagger for
-Turkish. Nevertheless, I expect my POS-tagger
-to work with a decent accuracy that is adequate to be a starting point for
-further development. Moreover, \citet{Korkut2015} began the development of a
-Turkish NLP library named ``Guguk'', implemented in the programming language Haskell. I intend
-to merge this project to Guguk, hence contributing towards
-the existence of a Turkish NLP-library in Haskell.
 
-Before proceeding, a few remarks on notational conventions: I use
-\texttt{HMM} (with the fixed-width font) to the datatype I'm using from the
-\texttt{Data.HMM} package and HMM (with regular font) to refer to the concept.
+The main motivation for this project comes from the lack of a POS tagger for
+Turkish exclusively implemented in the programming language Haskell. Moreover, I
+believe that implementing a POS tagger from scratch is beneficial in terms of
+gain experience with relevant statistical models through this process.
+
+Due to time and resource constraints, it is not plausible to build a
+state-of-the-art POS tagger for Turkish. Nevertheless, I expect my POS tagger to
+work with decent accuracy that is adequate to be a starting point for further
+development. Moreover, \citet{Korkut2015} began the development of a Turkish NLP
+library named ``Guguk'', implemented in the programming language Haskell. I
+intend to merge this project to Guguk, hence contributing towards the existence
+of a Haskell Turkish NLP library; I intend to continue development of this
+project as a part of Guguk.
+
+Before I proceed, let me give a few remarks on notational conventions I follow
+throughout this text: I use \texttt{HMM} (with the fixed-width font) to the
+datatype I'm using from the \texttt{Data.HMM} package and HMM (with regular
+font) to refer to the concept.
 
@@
 This proves an effective approach seeing that (1) Turkish morphology is
-exclusively based on suffixes (not prefixes) (2) morphology is central to the
-Turkish language so it is likely that any given word is formed through the
-agglutination of at least one morpheme. The tagger described has ``\%90.2
-accuracy in the best case''.
+exclusively based on suffixes, in contrast to to prefixes (2) morphology is
+central to the Turkish language so it is likely that any given word is formed
+through the agglutination of at least severals morphemes. The tagger described
+by \citet{dincer2008suffix} has ``\%90.2 accuracy in the best case''.
 
@@
 morphology, unlike most of the related work I've mentioned. There are definite
-problems with this approach. Fixing these would go beyond the scope of this
+problems with this approach; fixing these would go beyond the scope of this
 paper therefore I defer the integration of morphology for further development.
@@
 \section{Results}
-The resulting program I have implemented is based on bigram hidden markov
-models. To summarize it we are making use of three probabilities: (1) $P(t_i |
-t_{i-1})$ the probability of being at a state (i.e., a tag) given the previous
-state. This is computed\footnote{Although I use the word ``compute'', everything
-  is already pre-computed and the referred computation is merely a lookup. This
-  is discussed in \ref{par:tag}} by the function \texttt{transFn}. (2) $P(w_i | t_i)$ the
-probability of seeing a word having a given tag, computed by the function
-\texttt{outFn}. Finally, the probability of a given state being the start state
-of a sentence, computed by the function \texttt{outProbs}.
+
+The resulting program I have implemented is based on bigram hidden markov model.
+To summarize it, I am making use of three probabilities: (1) $P(t_i | t_{i-1})$
+the probability of being at a state (i.e., a tag) given the previous state. This
+is computed\footnote{Although I use the word ``compute'', everything is already
+  pre-computed when a call to this function is made and the referred computation
+  is merely a lookup. This is discussed in \ref{par:tag}} by the function
+\texttt{transFn}. (2) $P(w_i | t_i)$ the probability of seeing a word with a
+given tag, computed by the function \texttt{outFn}. Finally, the probability of
+a given state being the start state of a sentence, computed by the function
+\texttt{outProbs}.
 
 \subsection{A Brief Overview of the Program}
+
 The project consists of two executables: (1) \texttt{create-model} that reads
@@
 construct an \texttt{HMM}. Unfortunately, it is highly impractical---if not
-impossible---to directly serialize \texttt{HMM} since it has three fields
-containing functions
-\footnote{http://stackoverflow.com/questions/17785916/can-haskell-functions
+impossible---to directly serialize the \texttt{HMM} datatype since it has three
+fields containing functions \footnote{See
+  http://stackoverflow.com/questions/17785916/can-haskell-functions
   -be-serialized for further information.}. We consider this a fault in the
-design of the \texttt{Data.HMM} package and urge further development. Although
-this has almost no effect on performance, it obfuscates the code.
+design of the \texttt{Data.HMM} package and urge further development; this has
+almost no effect on performance, but it obfuscates the code.
 
@@
 \texttt{create-model} is intended be run only once after which \texttt{run-tag}
-can be used to make use of that data.
+can be used to make use of the serialized components of the HMM.
+
+I should note that it is not possible to give a complete account of the
+program's logic in this chapter although we explain the most salient components.
+The source code is in the \texttt{ayberkt-final_project} repository and it is
+well-commented. Should the reader have any concerns regarding implemenetation
+they could find it convenient to consult the source code.
 
 \subsubsection{\texttt{create-model}}
+
 The executable \texttt{create-model} is implemented in two files:
 \texttt{Parse.hs} and \texttt{Main.hs}. \texttt{Parse.hs} handles the parsing of
-the corpus.
+the treebank.
 
-The function \texttt{freqMap} takes as input a list of orderable elements
-(denoted with \texttt{Ord a}) and creates a \texttt{Map} containing, as value,
-the frequency of each element.
+The function \texttt{freqMap} found in \texttt{create/Main.hs} takes as input a
+list of orderable elements (denoted with \texttt{Ord a}) and creates a
+\texttt{Map} containing, as value, the number of ocurrence of each element.
 
@@
 This function is called several times with lists of values we want to
-\emph{count} such as tag bigrams, a given word with a certain tag et cetera.
-Pre-computing the counts is almost mandatory since frequencies (i.e., counts)
-will need to be accessed frequently when computing the probabilities. Once the
-necessary maps are generated, they are written to files in the directory
-\texttt{model}.
+\emph{count} (such as the list of tag bigrams, or the list of all tagged words)
+et cetera. Pre-computing the counts is almost mandatory since frequencies (i.e.,
+counts) will need to be accessed many times by the viterbi algorithm when
+finding the most probable path through the model. Once the necessary maps are
+generated, they are written to files in the directory \texttt{model}.
 
@@
 \label{par:tag}
+
 The executable \texttt{run-tag} is where the actual tagging happens. The
@@
 \end{lstlisting}
-
 This is a straightforward implementation of probability relying on a given count
 map. One thing to note is that it returns $\epsilon$ probability (i.e., the
-smalles value a \texttt{logFloat}) if a given element is not found.
+smallest value a \texttt{LogFloat}\footnote{A float in the log domain.} can have) if a given element is not found.
 
-\texttt{probability} when combined with \texttt{freqMap} forms an
-efficient method for implementing Bayes' law. When we are creating the functions
-required by \texttt{HMM} we are storing the map returned by a call to
-\texttt{freqMap} in the lexical closure of \texttt{probability} by making use
-Haskell's currying so that we can eloquently define the two crucial functions as
-follows:
+\texttt{probability} when combined with \texttt{freqMap} forms an efficient
+method for implementing dependent probability.
+
+When we are creating the functions required by \texttt{HMM} we are storing the
+map returned by a call to \texttt{freqMap} in the lexical closure of
+\texttt{probability} by making use Haskell's currying so that we can eloquently
+define the two crucial functions as follows:
 \begin{lstlisting}[
@@
 \end{lstlisting}
-The probability functions used to create \texttt{transFn} and \texttt{outFn} hold
-the count maps \texttt{tagBigramFreqs} and \texttt{taggedWordFreqs} in their
-lexical closure, respectively.
+The probability functions used to create \texttt{transFn} and \texttt{outFn}
+hold the count maps \texttt{tagBigramFreqs} and \texttt{taggedWordFreqs} in
+their lexical closure, respectively.
 
 Finally we declare a new \texttt{HMM} as follows:
-
 \newpage
@@
 
-This is a simple record (a record is like C struct in Haskell) containing the
-probability functions we defined.\footnote{For some reason, the author of the
-  \texttt{Data.HMM} library named the fields (i.e., attribute), that are of
-  function types, using names that end with \texttt{-Matrix} although they are
-  \emph{not} matrices. The reader should not be confused by this.}
+This is a simple record\footnote{A record is a datatype declaration in Haskell
+  similar to a struct in C.} containing the probability functions we
+defined.\footnote{For some reason, the author of the \texttt{Data.HMM} library
+  named the fields (i.e., attribute), that are of function types, using names
+  that end with \texttt{-Matrix} although they are \emph{not} matrices. The
+  reader should not be confused by this.}
 
 \subsection{Performance}
-Parsing the entire corpus and creating a model out of it takes 10.5 seconds on
-average. Tagging a sentence (which we are assuming to have 5 words on average)
-takes $~0.3$ seconds. The implementation of the \texttt{Data.HMM} datatype uses
-\texttt{Data.Array}
+Parsing the entire corpus and creating a model out of it takes $10.5$ seconds on
+average.\footnote{On my i5 MacBook} Tagging a sentence grows with $O(n)$
+complexity where $n$ is the number of words in a given sentence. Assuming a
+sentence has 5 words on average, this operation takes around $~0.3$ seconds.
 \footnote{https://hackage.haskell.org/package/array-0.5.1.0/docs/Data-Array.html}
-which is why it runs reasonably fast.
 
@@
 We have shown that it is possible---though not ideal---to build a POS tagger for
-Turkish that is based entirely on a bigram hidden markov model. Let us first
-look at the accuracy of this model and propose possible fixes that do not
-fundamentally change the structure of the model.
+Turkish---a canonical agglutinative language---that is based entirely on a
+bigram HMM. In this section I analyze the accuracy of the tagger and discuss
+simple improvements I can make, as well as its fundamental limitations fixing
+which would require a significant rewrite of the tagger. Let us first look at
+the accuracy of this model and propose possible fixes that do not fundamentally
+change the structure of the model.
 
@@
 \subsection{Limitations}
-There is a fundamental limitation to the POS tagger I built: Turkish morphology
-is essential to the overall completeness of Turkish.
+There exists a fundamental limitation to the described tagger: Turkish
+morphology is essential to the overall completeness of Turkish.
 
 In English we entertain the notion of \emph{syntactic correctness}, which is
-more or less a formalism of correct ordering of words. A speaker of English that
-has synactic competence in the language \citep{bachman1990} may know that a
+more or less a formalism of the correct ordering of words. A speaker of English
+that has synactic competence in the language \citep{bachman1990} may know that a
 sentence is well-formed, relying on his knowledge of syntactic rules. However,

### 2015-12-09 paper/paper.tex
Write on further research

@@
 The project consists of two executables: (1) \texttt{create-model} that reads
-the data from the corpus and parses it into a form from which we can easily
+the data from the corpus and parses it into a form from which one can easily
 construct an \texttt{HMM}. Unfortunately, it is highly impractical---if not
@@
   http://stackoverflow.com/questions/17785916/can-haskell-functions
-  -be-serialized for further information.}. We consider this a fault in the
+  -be-serialized for further information.}. I consider this a fault in the
 design of the \texttt{Data.HMM} package and urge further development; this has
@@
 I should note that it is not possible to give a complete account of the
-program's logic in this chapter although we explain the most salient components.
-The source code is in the \texttt{ayberkt-final_project} repository and it is
+program's logic in this chapter although I explain the most salient components.
+The source code is in the \texttt{ayberkt-final\_project} repository and it is
 well-commented. Should the reader have any concerns regarding implemenetation
@@
 
-This function is called several times with lists of values we want to
-\emph{count} (such as the list of tag bigrams, or the list of all tagged words)
+This function is called several times with lists of values designate to be
+\emph{counted} (such as the list of tag bigrams, or the list of all tagged words)
 et cetera. Pre-computing the counts is almost mandatory since frequencies (i.e.,
@@
 The executable \texttt{run-tag} is where the actual tagging happens. The
-frequencies we have pre-computed in \texttt{create-model} are passed in to the
-probability function which we give below.
+pre-computed frequencies  in \texttt{create-model} are passed in to the
+probability function which I give below.
 \begin{lstlisting}[
@@
 
-When we are creating the functions required by \texttt{HMM} we are storing the
+When I am creating the functions required by \texttt{HMM} I am storing the
 map returned by a call to \texttt{freqMap} in the lexical closure of
-\texttt{probability} by making use Haskell's currying so that we can eloquently
-define the two crucial functions as follows:
+\texttt{probability} by making use Haskell's currying so that the functions can
+be eloquently defined as follows:
 \begin{lstlisting}[
@@
 
-Finally we declare a new \texttt{HMM} as follows:
-\newpage
+Finally a new \texttt{HMM} is declared as follows:
 \begin{lstlisting}[
@@
 This is a simple record\footnote{A record is a datatype declaration in Haskell
-  similar to a struct in C.} containing the probability functions we
+  similar to a struct in C.} containing the probability functions I have
 defined.\footnote{For some reason, the author of the \texttt{Data.HMM} library
@@
 
-We have shown that it is possible---though not ideal---to build a POS tagger for
+I have shown that it is possible---though not ideal---to build a POS tagger for
 Turkish---a canonical agglutinative language---that is based entirely on a
@@
 
-We have tested the POS tagger with 6075 words from a small portion of the corpus
+I have tested the POS tagger with 6075 words from a small portion of the corpus
 that I reserved for testing (i.e., did not use for tranining the HMM). The POS
@@
   underscore. Since there is not an easy way to detect whether two sequential
-  words are reduplications or not, we are giving duplications in an incorrect
-  format to the HMM.  We predict that (based on informal tests with the model)
+  words are reduplications or not, I am giving duplications in an incorrect
+  format to the HMM.  I predict that (based on informal tests with the model)
   that correction of this problem would significantly increase the accuracy.
@@
   this decreases the accuracy of the HMM: instead of a tag $t$ being followed by
-  a tag $t'$, we get a lot of $t, <clause>, t'$ which obscures the word order we
-  are interested in. Since we may always assign the \textsc{Punc} tag to a comma
-  with full certainty, we may simply ignore commas as far as probability is
-  concerned and tag them with \textsc{Punc}.
+  a tag $t'$, there are a lot of $t, <clause>, t'$ which obscures the word order
+  we are interested in. Since it is possible to always assign the \textsc{Punc}
+  tag to a comma with full certainty, I may simply ignore commas as far as
+  probability is concerned and tag them with \textsc{Punc}.
 
@@
 
-In English we entertain the notion of \emph{syntactic correctness}, which is
-more or less a formalism of the correct ordering of words. A speaker of English
-that has synactic competence in the language \citep{bachman1990} may know that a
+In English, there exists a notion of \emph{syntactic correctness}, which is more
+or less a formalism of the correct ordering of words. A speaker of English that
+has synactic competence in the language \citep{bachman1990} may know that a
 sentence is well-formed, relying on his knowledge of syntactic rules. However,
@@
 is why computational processing of Turkish can never fully escape its
-morphology.
+morphology. In the case of the described tagger, the \texttt{outProbFn} function
+becomes increasingly unreliable as the morphological complexity of a word
+increases; for words with more than six-seven suffixes, it is unlikely that the
+word has been seen before.
 
@@
 suffixes, that change their form as not to upset the major vowel harmony in a
-given word. We may say with probably more than $90\%$ probability that a given
-word has a certain POS looking whether its ending satisfies regular expressions.
+given word. It almost \emph{certain} that a given word has a certain POS looking
+whether its ending satisfies regular expressions.
 
-\bibliographystyle{apalike}
+As a more paradigmatically challenging remark, I suggest the possibility of
+tackling Turkish from a Distributed Morphology \citep{halle1993distributed}
+approach. With a theory that views syntax and morphology as instances of the
+same generative engine, it might be possible to expose to true nature of
+agglutinative languages hence dramatically increasing the accuracy and reducing
+the complexity such a program. \bibliographystyle{apalike}
 \bibliography{your_bibliography}

### 2015-12-09 paper/paper.tex
Fix typos

@@
 %include agda.fmt
-\maketitle
 
@@
 believe that implementing a POS tagger from scratch is beneficial in terms of
-gain experience with relevant statistical models through this process.
+gaining experience with relevant statistical models.
 
 Due to time and resource constraints, it is not plausible to build a
-state-of-the-art POS tagger for Turkish. Nevertheless, I expect my POS tagger to
-work with decent accuracy that is adequate to be a starting point for further
+state-of-the-art POS tagger for Turkish. Nevertheless, I expect the POS tagger
+to work with decent accuracy that is adequate to be a starting point for further
 development. Moreover, \citet{Korkut2015} began the development of a Turkish NLP
-library named ``Guguk'', implemented in the programming language Haskell. I
-intend to merge this project to Guguk, hence contributing towards the existence
-of a Haskell Turkish NLP library; I intend to continue development of this
-project as a part of Guguk.
+library named ``Guguk'', implemented in Haskell. I intend to merge this project
+to Guguk, hence contributing towards the existence of a Haskell Turkish NLP
+library; I also plan to continue working on this project as a part of Guguk.
 
-Before I proceed, let me give a few remarks on notational conventions I follow
-throughout this text: I use \texttt{HMM} (with the fixed-width font) to the
-datatype I'm using from the \texttt{Data.HMM} package and HMM (with regular
-font) to refer to the concept.
+Before I proceed, let me make a few remarks on the notational conventions I
+follow throughout this text: I use \texttt{HMM} (with the fixed-width font) to
+refer to the datatype I'm using from the \texttt{Data.HMM} package and HMM (with
+regular font) to refer to the concept.
 
@@
 
-% Briefly survey the most salient prior work that relates to the problem you wish
-% to solve. This section should cite relevant sources. This section should be at
-% least one to two paragraphs in length.
+% Briefly survey the most salient prior work that relates to the problem you
+% wish to solve. This section should cite relevant sources. This section should
+% be at least one to two paragraphs in length.
 \cite{dincer2008suffix} built a POS tagger for Turkish using HMMs on suffixes.
@@
 impossible---to directly serialize the \texttt{HMM} datatype since it has three
-fields containing functions \footnote{See
+fields containing functions.\footnote{See
   http://stackoverflow.com/questions/17785916/can-haskell-functions
-  -be-serialized for further information.}. I consider this a fault in the
+  -be-serialized for further information.} I consider this a fault in the
 design of the \texttt{Data.HMM} package and urge further development; this has
@@
 Since processing the corpus is a resource-demanding operation,
-\texttt{create-model} is intended be run only once after which \texttt{run-tag}
-can be used to make use of the serialized components of the HMM.
+\texttt{create-model} is intended to be run only once after which
+\texttt{run-tag} can be used to make use of the serialized components of the
+HMM.
 
@@
 program's logic in this chapter although I explain the most salient components.
-The source code is in the \texttt{ayberkt-final\_project} repository and it is
-well-commented. Should the reader have any concerns regarding implemenetation
-they could find it convenient to consult the source code.
+The source code is in the \texttt{ayberkt-final\_project} repository and is
+well-commented. Should the reader have any concerns regarding the specifics of
+the implementation they could find it convenient to consult the source code.
 
@@
 list of orderable elements (denoted with \texttt{Ord a}) and creates a
-\texttt{Map} containing, as value, the number of ocurrence of each element.
+\texttt{Map} containing, as value, the number of occurrences of each element.
 
@@
 ]
-freqMap :: Ord a $\Rightarrow$ [a] $\rightarrow$ M.Map a Int
+freqMap :: Ord a $\Rightarrow$ [a] $\rightarrow$ Map a Int
 freqMap elems = populate empty elems
-  where populate :: Ord a Map a Int $\rightarrow$ [a] $\rightarrow$ M.Map a Int
+  where populate :: Ord a Map a Int $\rightarrow$ [a] $\rightarrow$ Map a Int
          populate m [] = m
-         populate m (x:xs) = let freq = (M.findWithDefault 0 x m) :: Int
-                              in populate (M.insert x (freq + 1) m) xs
+         populate m (x:xs) = let freq = (findWithDefault 0 x m) :: Int
+                              in populate (insert x (freq + 1) m) xs
 \end{lstlisting}
 
-This function is called several times with lists of values designate to be
+This function is called several times with lists of values designated to be
 \emph{counted} (such as the list of tag bigrams, or the list of all tagged words)
@@
 \texttt{probability} when combined with \texttt{freqMap} forms an efficient
-method for implementing dependent probability.
+method for computing dependent probability.
 
